Autotrophic growth of E. coli is achieved by a small number of genetic changes

Version 1

Abstract

Abstract

Synthetic autotrophy is a promising avenue to sustainable bioproduction from CO2. Here, we use iterative laboratory evolution to generate genetically diverse autotrophic strains. We identify that just three mutations are sufficient for E. coli to grow autotrophically, when introduced alongside non-native energy (formate dehydrogenase) and carbon-fixing (Rubisco, phosphoribulokinase, carbonic anhydrase) modules. The mutated genes are involved in glycolysis (pgi), central-carbon regulation (crp), and RNA transcription (rpoB). The pgi mutation reduces the enzyme activity, thereby stabilising the carbon-fixing cycle by capping a major branching flux. The other two mutations increase the ratio of NADH/NAD+ - the cycle’s electron-donor. This study demonstrates the malleability of metabolism and evolution’s capacity to switch trophic modes on laboratory time-scales and could facilitate transforming other heterotrophic organisms into autotrophs.

Introduction

Anthropogenic interference and intensive burning of fossil fuels releases large amounts of sequestered carbon in the form of CO2, a main driver of climate change (1, 2). Direct air carbon capture, by any means, is urgently needed to help reduce emissions. Here, we focus on biology based carbon fixation as a potential avenue for CO2 capture which has been shown to be a potentially efficient production platform(3–5). Specifically, use microbes for CO2 valorization and produce a diverse repertoire of products including food, fuel, bio-plastics or other commodities(6–9). Natural autotrophs already metabolise CO2 efficiently, yet most are challenging to cultivate and manipulate genetically compared to heterotrophic model organisms like E. coli. Here we explore an emerging alternative approach – converting heterotrophic model organisms into tractable autotrophs for bio-production(7, 10–12). In addition to promising engineering applications, studying the transition from heterotrophic to autotrophic metabolism can uncover fundamental principles about the structure and regulation of carbon fixation metabolism(11–16) .

Recently, with the increasing interest in platforms for sustainable production, synthetic biologists have been able to successfully manipulate different organisms to enable major metabolic transitions (11, 12, 17–22). These efforts resulted in the successful introduction of non-native C1 utilisation pathways, and even included successful transitions from heterotrophic into completely autotrophic metabolisms in bacteria (12) and yeast (11). These previous studies introduced the reductive pentose phosphate (rPP) cycle, the most prominent carbon fixation pathway in nature also known as the Calvin-Benson cycle, in Escherichia coli (E. coli) and Komagataella phaffii (previously Pichia Pastoris). Indeed, we showed that E. coli was able to utilise the rPP cycle for the synthesis of all biomass carbon - thus converting it from a heterotroph to an autotroph(12); However, this transition required continuous culture in selective conditions for several months (adaptive lab evolution) and resulted in many uninterpretable mutations, exposing a large knowledge gap.

Using adaptive lab evolution presents a challenge as it is hard to separate the essential changes allowing a phenotype from the overall set of accumulated mutations. Here we aimed at characterising the landscape of mutations required for autotrophic growth. Studying mutations that arise in multiple lineages can help us better understand the biological basis for the heterotroph-to-autotroph transition. To achieve this goal, we created a pipeline that enabled us to distinguish mutations essential for autotrophy from low-frequency mutations that promoted the adaptive evolution experiments and determine a compact essential set of changes permitting autotrophy in E. coli. We then proceeded with further analyses to shed light on the mechanism behind this compact essential set of mutations, showing the phenotype included a change in the activity of a central carbon metabolism enzyme and in intracellular changes of the non-native carbon fixation cycle’s metabolite and co-factors pools. We continue by speculating about the role of these metabolic changes.

Results

A compact set of mutations that enable autotrophic growth was identified using rational design, iterative lab evolution and genetic engineering

Adaptive lab evolution is an effective tool that enables integration of non-native metabolic pathways. In order to harness the power of adaptive lab evolution, a selection system must be implemented, to direct the evolution towards the desired function. In our case, selection for the activity of the non-native rPP cycle. To ensure that some level of carbon fixation by rubisco becomes essential for growth, we knocked out phosphofructokinase (pfkA and pfkB), and 6-phosphate-1-dehydrogenase (zwf) which creates a stoichiometric imbalance and growth arrest when growing on five carbon sugars such as xylose. This imbalance could be rescued by a metabolic bypass - the phosphorylation of ribulose-5-P by Prk and the subsequent carboxylation by rubisco, thereby coupling growth to the heterologously expressed rPP enzymes. This setup allowed us to evolve E. coli, a natural heterotroph, into an autotroph, using a non-native rPP cycle to utilise CO2 as a sole carbon-source and Formate-dehydrogenase to oxidise formate for energy (12). However, using adaptive lab evolution also presents a challenge -once a new phenotype has been achieved, it is hard to separate the essential changes allowing for its appearance from the overall set of accumulated mutations. In many cases, mutations occur on the background of highly adaptive ones, leading to a spread of neutral “hitchhiker” mutations in the population. In other cases, mutations that appeared at the beginning of the evolution are no longer needed for the final phenotype. Our goal was to find a set of mutations that, if introduced into a heterotrophic wild type E. coli, would result in the usage of the non-native carbon fixation and energy modules. This would enable autotrophic growth based solely on genetic engineering without the need for further evolutionary adaptation. Isolating the essential genetic changes would facilitate studying the mechanism behind this metabolic transition. To do so, we used a workflow that included three main stages as shown in Fig. 1. A) rational design -introduction of required heterologous genes and metabolic knockouts to enforce Rubisco-dependent growth; B) adaptive lab evolution -revealing mutation candidates by growing the designed strain in autotrophic-selecting conditions over many generations until an autotrophic phenotype is obtained; C) evolution-inspired genetic engineering -introducing the most promising mutations revealed in stage B into the designed strain and testing it for autotrophic growth. Stages B and C were repeated until no further evolution was required, i.e. the designed strain in stage C had an autotrophic phenotype. To find the most promising mutations in stage B, we applied at least one of two criteria: mutations in genes that occur repeatedly in different autotrophic lineages (but are uncommon in other adaptive lab evolution experiments) or mutations that when reverted, result in a loss of phenotype. We discovered these “consensus” mutations by sequencing isolated autotrophic clones, obtained from repeated adaptive lab evolution experiments -similar to the conditions we used in Gleizer et al. 2019 (Fig. 1; steps B1, B2). In practice, we inoculated the strain denoted as Ancestor to a chemostat with limited xylose and excess formate. After ≈3 months (≈60 chemostat generations) we isolated from the chemostat a clone able to grow under autotrophic conditions (Methods), and compared its mutation set to those of two previously evolved autotrophic clones(12). The intersection of the three clones contained 4 genes that were mutated in all autotrophic isolated clones: the central carbon metabolism enzyme phosphoglucoisomerase (pgi), the beta subunit of the RNA polymerase (rpoB), poly A polymerase (pcnB) and the MalT DNA-binding transcriptional activator (malT) (Fig. 1, step B1). rpoB, pcnB and malT are genes that are commonly mutated in adaptive lab evolution experiments(23), suggesting that they have little relevance for autotrophic growth. Therefore, we chose to focus on the pgi mutation. We introduced the H386Y pgi mutation into the rationally designed ancestor (Fig. 1, step C1), and the engineered strain was tested for growth in autotrophic conditions (methods). No growth was observed in those conditions, which meant that additional mutations were necessary in order to achieve autotrophy. Genomic sequencing revealed that during the genetic manipulation for introducing the pgi mutation, another mutation in rpoB (A1245V) appeared in the genome as well. In parallel, reverting a variety of RNA polymerase mutations in other evolved strains back to their wild-type allele, showed that mutations in RNA polymerase are in fact essential to the phenotype. Therefore, despite the fact that it was unintentional, we decided to leave the mutation in the genome and move on with this strain. We then used this strain for another round of evolution (Fig. 1, step B2). Since we started the evolution experiment with a strain mutated in both pgi & rpoB (denoted “ancestor 2.0”) we expected that the desired phenotype appears in fewer generations, and therefore fewer mutations. Indeed, on two separate experiments, the autotrophic phenotype appeared at a considerably shorter period of time when harbouring both mutations. In both cases, it took roughly one month until autotrophic growth was observed, compared to >90 days in previous attempts. Even though these strains (Fig. 1, “Evolved 4” and “Evolved 5”) both had multiple mutations, only one mutation was shared between these independent evolution experiments - a non-synonymous mutation in the cAMP receptor protein, crp (H22N). When we introduced this mutation to the “ancestor 2.0” background in the evolution-inspired genetic engineering stage (Fig. 1, step C2), the strain was immediately able to grow under autotrophic conditions. Therefore, three mutations (pgi*, crp* and rpoB*) were sufficient to facilitate the autotrophic phenotype. Since the RNA polymerase mutation was introduced into this strain (rpoB A1245V) as a byproduct of genetic engineering and never observed in the adaptive evolution experiments, we tested its essentiality to the autotrophic phenotype. We found that a reversion of rpoB A1245V back to the wild-type allele causes a loss of the autotrophic phenotype. We denote the final engineered autotrophic E. coli strain, with only 3 mutations on top of the heterotrophic ancestor strain, the “compact autotrophic E. coli”.

Autotrophic phenotype achieved by introducing 3 mutations on top of a rationally designed ancestor.

(A) We rationally designed the “wild-type” E. coli background strain (BW25113, depicted as a black bacterium) by introducing 4 enzymes: RuBisCO (cbbM), phosphoribulokinase (prkA), carbonic anhydrase (CA) andformate dehydrogenase (fdh), and by 3 genomic knockouts: glucose-6-phosphate dehydrogenase (zwf) and phosphofructokinase A&B (pfkAB). We denote the resulting strain as “ancestor” (brown bacterium). (B) We tested the resulting strain for autotrophic growth and, since it didn’t grow, we used it for adaptive lab evolution in xylose-limited chemostats -i.e., conditions that select for higher carbon fixation flux. Altogether, we were able to isolate three evolved clones (two distinct strains from one chemostat experiment, Evolved I & II, and another strain from a second experiment, Evolved III, step B1). (C) Out of four consensus mutations that we identified, two -rpoB and pgi -were incorporated into the ancestor to get “ancestor 2.0” (light blue bacterium, step C1). Once more, we tested for autotrophic growth and were unsuccessful. Therefore, we initiated another round of adaptive lab evolution experiments using ancestor 2.0 growing in two xylose limited chemostats and isolated two evolved clones (one from each chemostat, step B2). The two clones shared a single consensus mutation, in crp. We thus created a new strain from ancestor 2.0 by introducing the crp mutation to it (blue bacterium, step C2). This strain could grow autotrophically and thus we achieved a compact autotrophic strain.

We verified the genotype using whole genome sequencing (methods). The results included the rationally designed knockouts (ΔpfkA, ΔpfkB, Δzwf), heterologous plasmids (energy and carbon-fixing modules) and the three introduced mutations (pgi*, crp* and rpoB*). During the genetic engineering process, two additional mutations occurred unintentionally: in the genes uhpT and yejG. uhpT is a hexose transporter and is unlikely to have any effect on the autotrophic phenotype, especially because the mutation was an early stop codon (Q7*) and likely a loss of function. As the function of yejG is unknown, we wanted to ensure that it is not essential to the autotrophic phenotype. Therefore, we reverted the mutation back to its wild-type allele and found that indeed the cells were still autotrophic.

Thus, we constructed a compact strain, which had three essential mutations (pgi, crp & rpoB). When comparing this compact strain to a previously characterised autotrophic strain(12), it was nearly identical in terms of growth rate, lag time and yield (Fig. 2A). Using 13C labelling, we verified that all the biomass carbon stems from CO2 (Fig. 2B). We termed the three introduced mutations (pgi, crp & rpoB) together with the carbon fixation machinery (cbbM, prkA & CA) and the energy module (fdh) the “autotrophic enabling gene set”.

Validation and characterisation of the autotrophic phenotype.

(A) left: growth curve of an isolated evolved clone vs the engineered compact strain in liquid M9 minimal media supplemented with 45mM sodium formate and sparged with a gas mixture of 10% CO2, 5% oxygen, and 85% nitrogen. right: calculated growth rate using linear regression of the engineered compact strain (light blue) and evolved strain (orange), the calculated doubling time at the given conditions is about 24h (μ ≈ 0.03h-1). Growth was carried out in triplicates (n=3) in a Spark plate reader with gas control, dark grey area corresponds to standard deviation (150ul culture + 50ul mineral oil, to prevent evaporation). (B) The fractional contribution of 13CO2 to various protein-bound amino acids of the compact autotrophic strain after ≈4 doublings on 13CO2 and 13C labeled formate (light blue, mean of n=3) reached the expected 13C labeling fraction of the biosynthesized amino acids. When grown in naturally labeled CO2 and 13C labeled formate (dark blue, n=1, technical triplicate) the 13C content dropped to close to natural abundance. Experiments with 13CO2 as the substrate were carried out in air-tight (i.e., sealed) growth vessels.

Introduction of autotrophy-enabling gene set into wild-type background doesn’t require bypass-prevention for growth

As described above, in order to select for autotrophic growth, the native E. coli was metabolically re-wired by 3 auxiliary genomic knockouts (ΔpfkA, ΔpfkB, Δzwf). These knockouts created dependency on carboxylation of CO2 by Rubisco for growth even when consuming pentose sugars, which were supplemented during the chemostat evolution (methods) (12, 13). This should direct the evolution towards increased usage of the synthetic rPP cycle. Once autotrophic growth was achieved, and the enabling gene set was found, we wanted to test if these knockouts (pfkA pfkB and zwf) are still required for the phenotype, now that the evolutionary selection process was completed. We tested this by introducing the set of autotrophy-enabling genes into a wild-type (BW25113) E. coli background (methods, Fig. 3). After doing so, we discovered an unintentional genomic mutation, in the ribosomal 16S subunit (rrsA) that was introduced during the genomic manipulation process. The engineered E. coli strain was able to grow autotrophically, albeit reaching 2-fold less than the final OD of the rationally designed compact strain still harbouring the auxiliary knockouts. This indicates that the “autotrophic enabling gene set” is sufficient to achieve the trophic mode change and that the knockouts are not essential for the final autotrophic growth. As discussed below, this indicates that in some cases metabolic rewiring could be used as a temporary “metabolic scaffold” and could be removed from the genotype once the desired phenotype is achieved. We proceeded by conducting various comparative assays, including metabolomics and physiology experiments to test the effect of the three mutations on the cell.

Auxiliary knockouts are not required for final phenotype.

We transformed a wild-type BW25113 E. coli strain (black bacteria) with the carboxylating and energy module plasmids (grey circles with coloured gene annotations), and inserted 3 auxiliary genomic knockouts (red octagon) to rewire metabolism toward carboxylating dependency. This strain was used for iterative evolution experiments in order to generate diverse autotrophic strains and reveal mutations candidates for rational design, i.e. consensus mutations. The identified autotrophic enabling mutations (thin grey circle) were then introduced in a wild type strain expressing the heterologous plasmids but without the auxiliary knockouts, the final strain was able to grow in autotrophic conditions. Dashed lines represent gene/mutation introduction.

The compact autotrophic E. coli contains only three genomic mutations and provides an opportunity to derive design principles that might serve as guidelines for future engineering efforts that use the rPP cycle. Therefore, we designed experiments that aimed to identify the relevant phenotypes of each of these mutations. Because the mutations were essential for the autotrophic growth, we could not use autotrophic conditions to compare them to a strain with wild-type alleles. Therefore, we chose the most suitable combinations of strains and conditions that could isolate the effect of the pertinent mutation.

Modulation of a metabolic branch-point activity increased the concentration of rPP metabolites

Metabolic pathways that regenerate and synthesise more of their own metabolites are referred to as autocatalytic cycles. Within the autotrophic E. coli the rPP cycle is autocatalytic. Due to the inherent positive feedback mechanism, autocatalytic cycles tend to be unstable, and therefore the fluxes of entry and exit points (bifurcation/branch points) need to be balanced (14). Such tuning is needed since any disruption of the balance between cycling and branching fluxes could result in the depletion or alternatively toxic accumulation of intermediate metabolites and cycle arrest. We previously predicted that mutations in branch points will be needed to stabilise the steady state flux within the rPP cycle (14). In line with this prediction we find that the mutation in Phosphoglucoisomerase (Pgi), a key branch point in the rPP cycle, follows this design principle. Pgi consumes fructose-6-phosphate (F6P), one of the cycle intermediates, converting it into glucose-6-phosphate (G6P), a precursor for cell membrane biosynthesis. Thus, Pgi regulates flux out of the autocatalytic cycle (Fig. 4A). In the different adaptive lab evolution experiments, we found three distinct mutations in the pgi gene. The first, H386Y, is a non-synonymous mutation occurring in one of the catalytic residues in the active site, and is part of the autotrophic enabling gene set. The second was a complete knockout, a 22KB chromosomal deletion, including also 16 other genes. The third was an early stop codon E72*. These observations led us to suspect the H386Y mutation in pgi decreases or even completely eliminates the activity of the enzyme. To test this, the kinetic rates of the isomerization of F6P to G6P of two purified Pgi enzymes, wild type Pgi and Pgi H386Y, were determined using a spectroscopic coupled assay, where G6P was coupled to NADP+ reduction (24) (Fig. 4B; see also methods). The measured kcat of the wild type Pgi was ≈130 [s-1] (±26) which is inline with previous computational predictions of a kcat of ≈200 [s-1] (25). The mutated Pgi showed weak activity with a kcat of ≈1.5 [s-1](±1). Therefore the pgi mutation is in line with the original expectation that branching point regulation is necessary, and the nature of the observed mutations in pgi supports the notion that their role is to reduce the efflux from the non-native autocatalytic cycle. Pgi is part of the glycolysis/gluconeogenesis, where significant flux is required in wild type E. coli. Accordingly, wild type Pgi catalyses a reaction that is causing a strong efflux of F6P from the cycle, diminishing the regeneration of ribulose-1,5-bisphosphate (RuBP) which is usually not needed in the native metabolism but is essential for the rPP autocatalytic activity. In the mutated version of Pgi the efflux capacity is diminished, which can stabilise the cycle (Fig. 4C). Following the same logic we can expect an increase in the F6P substrate metabolite pool. Therefore, we measured intracellular sugar-phosphates in the pgi mutant strain and a strain with a wild-type pgi allele by liquid chromatography-tandem mass spectrometry (LC-MS/MS). For this comparison, we used the rubisco-dependent ancestor strain as the genetic background. We found that the ratio of F6P to G6P was about 3 times higher in the pgi mutant strain relative to the wild-type (Fig. 4D). Furthermore, the pgi mutant had higher levels of metabolites within the rPP cycle (Fig. 4D), confirming the stabilising function of the mutation. These results lead us to conclude that the recurring mutations in the pgi gene are mediating the integration of the non-native carbon fixation genes into a stable autocatalytic cycle. This is achieved by reducing the ratio of the efflux from F6P to G6P relative to the flux in the cycle, thereby redirecting it towards the rPP cycle and the regeneration of RuBP.

Pgi mutation is required to partition flux towards regeneration of autocatalytic cycle substrate.

(A) Metabolic scheme of rPP cycle components when growing on xylose. (B) In vitro spectrophotometric coupled assay determined that the rate of the isomerization of fructose-6-phosphate (F6P) to glucose-6-phosphate (G6P) is ≈100-fold lower for purified PgiH386Y (≈1.5 [s-1]) compared to Pgiwild-type (≈130 [s-1]), n=3. Noise level was determined by negative control samples containing a non-Pgi enzyme. (C) Left: The wild type pgi, competes for its substrate F6P with the autocatalytic cycle, resulting in low F6P pool and low regeneration rate of ribulose-5-phosphate (Ru5P), which means that the rubisco-dependent pathway requires constant xylose supply. Right: The mutated Pgi (green) has reduced activity, which increases the ratio between the flux in the cycle and efflux which is needed for a stable regenerative flux towards Ru5P (via the pentose-phosphate-pathway), thus enabling an autotrophic cycle. (D) Left: Measured relative intracellular ratio of F6P and G6P of the ancestor background harbouring a PgiH386Y mutation compared to the ancestor with a wild-type Pgi, both grown in rubisco dependent conditions, n=6 cultures p-value < 0.05. Right: Relative intracellular abundance of pentose-phosphate-pathway metabolites -sedoheptulose-7P (S7P) and total pool of Pentose-phosphates (Ribulose-5P, Ribose-5P, Ribose-1P, and Xylulose-5P -denoted P5P) in a PgiH386Y strain versus the Pgiwild-type strain, both growing in a rubisco-dependent manner, n = 3 cultures.

Mutations in regulatory genes lead to increased availability of the carbon fixation cycle electron donor NADH

Along with the branch-point mutation in pgi, the compact autotrophic E. coli has two transcription-associated mutations, one in the gene crp, a global metabolism regulator and one in rpoB, a component of the transcriptional machinery. Since Crp and RpoB are known to physically interact in the cell (26–28), we address them as one unit, as it is hard to decouple the effect of one from the other. Mutations in crp and rpoB are known to affect growth at different conditions such as types of media or carbon sources (28–31). Therefore we tested whether this was also the case for the conditions under which the cells evolved in the chemostat. We performed a growth experiment of the double mutant (crp H22N, rpoB A1245V) expressing fdh in a minimal medium containing xylose and formate. Under such conditions, the strain harbouring both mutations and the fdh exhibited >25% increase in the final OD compared to the wild type expressing fdh (Fig. 5A&B). At the same time they also showed a longer (>10 hours) lag time. Notably, when grown solely on xylose, in the absence of formate, the increase in yield was not observed, and the extended lag was less pronounced (Fig. 5A).

mutations in rpoB and crp increase yield and intracellular NADH/NAD+ levels in fdh-expressing E. coli in the presence of formate.

Growth experiment of BW25113 wild-type (grey) compared to a crp H22N rpoB A1245V mutant (orange). Both strains express fdh. (A) strains were grown in 1g/L xylose and 40mM formate (solid line) or in the absence of formate (dashed line). The experiment was executed in a 96 well plate in 10% CO2 atmosphere, n=6 repeats. Lines represent the mean, light-grey background represents the standard deviation. (B) Maximal OD600 of the wild-type (grey) and mutant (orange) strains grown in 1g/L xylose and 40mM formate or in the absence of formate. Bar heights represent averages (n=6) of the median of the top 10 OD measurements of each replicate. Error bars represent standard error. (C) Intracellular NADH/NAD+ ratio of the wild-type (grey) was compared to the mutant (orange). The strains were grown in 2g/L xylose and 30mM formate, or in the absence of formate. The y-values are NADH/NAD+ ratios as fold-changes relative to wild-type without formate. Boxes represent 25-75 percentile ranges and dark lines represent median values. All data points are depicted without removing outliers.

Since NADH is the electron donor for carbon fixation, and a product of formate consumption via Fdh, we used metabolomics to measure the ratio of NADH/NAD+ in these strains in similar conditions (Methods). All of the measured ratios were normalised to the wild-type (with fdh) growing in the absence of formate. The addition of 30mM formate, which can serve as an electron donor via oxidation to CO2 by Fdh, increased NADH/NAD+ ratio by >20-fold in the double mutant, significantly higher than the ≈10-fold increase in the wild-type strain (p-value <0.05). No significant difference was observed between the strains in the absence of formate (Fig. 5C). NADH could affect growth in several ways, as discussed below. The specific reason for the increase in final OD or lag time remains inconclusive at this point. We also note that mutations in global transcriptional regulators like Crp have many effects on the autotrophic E. coli(32) which await future exploration.

Discussion

In this study, we demonstrated how iterative evolution and engineering were able to narrow down the genetic mutations required for autotrophy in E. coli from tens of mutations - to only three mutated genes. The three mutated genes -pgi, crp and rpoB -serve as a compact set, which when introduced into a wild-type background and supplemented with the carbon fixing and energy modules, achieve an autotrophic phenotype. By metabolic and physiological characterisation we were able to distil some of the effects of these mutations and understand their role in facilitating the autotrophic phenotype. We suggest that the two main effects achieved by the autotrophy-enabling mutations are: (i) integrating the non-native carbon fixation cycle by enhancing flux within the cycle over outgoing flux; and (ii) increasing the NADH/NAD+ ratio, which could have both thermodynamic and kinetic benefits.

The mutation in pgi affects the activity of a gene that encodes a phosphoglucoisomerase - a metabolic enzyme that catalyses a reaction that branches out of the non-native rPP cycle. The enzyme converts F6P to G6P, with the former being a cycle intermediate, and the latter a precursor required for membrane biosynthesis. In a previous study, this mutation was found to be essential for the integration of the rPP autocatalytic cycle and was part of a minimal set for hemi-autotrophic growth(16). As our results suggest, the mutated enzyme shows a significant reduction in activity, which decreases the efflux from the rPP cycle to biosynthesis. We hypothesised that its role is to generate a new, higher, steady-state concentration for the rPP cycle metabolites, thus stabilising the cycle. This was supported by our observation that introducing a pgi mutation into a metabolically rewired Rubisco-dependent strain resulted in increased concentration of rPP cycle intermediates. This exemplifies the key role branch-points play in autocatalytic cycles(12–14, 16), making them a leading target for rational attempts to integrate non-native autocatalytic cycles into metabolism.

The other two mutations, crp and rpoB, encode for proteins that interact with one another. We therefore addressed them simultaneously, finding that these mutations lead to a significant increase in NADH/NAD+ ratio -essential cofactors for the electron transfer reaction of the non-native rPP cycle (catalysed by the D-glyceraldehyde-3P dehydrogenase GAPDH). Under glycolytic conditions, the net reaction flows in the direction of D-glyceraldehyde-3P oxidation and, therefore, needs to be reversed for the rPP cycle. Increased levels of NADH, the reaction product in glycolytic conditions, facilitates such reversal, by tilting the thermodynamics in the opposite direction. Moreover, this increased ratio could offer a solution not only to a thermodynamic requirement, but also to a kinetic one. Since the amount of flux is a function of both the enzyme and substrate concentration - if we assume the enzyme concentration has not changed, increasing the substrate pool, as the ratio suggests, would result in increased flux. This result is in line with previous models, predicting that high cofactor levels are required for a stable metabolic state of the rPP cycle(33) and offers an avenue for future rational designs. Since both crp and rpoB affect gene expression globally, the role they play in the autotrophic phenotype likely extends beyond controlling the NADH level(32). In the future, this could be explored with various methods, such as proteomics, transcriptomics, ChIP-seq, or other assays for transcription factor activity. We speculate that such regulatory mutations are to be expected in adaptive laboratory evolution experiments. Especially during selection for a major shift (novelty) in the phenotype. A major shift will often necessitate a modulation of the activity of many cellular components. Lab evolution is limited by population size and number of generations, which means it can only explore a small part of the genotypic space. Therefore, sets of three or more mutations, where each mutation on its own does not confer an advantage, have a negligible probability to occur. On the other hand, a single mutation in a global regulator could concurrently perturb the activity of hundreds of components and might be the only viable solution to capture several required changes simultaneously.

We also show here how metabolic rewiring can serve as a “metabolic scaffold” to direct lab evolution. Since scaffold mutations are put in place for the purposes of selection, they can be removed once the goal is achieved. A common limitation of lab evolution is its tendency to follow evolutionary trajectories which increase fitness by deactivating or losing the heterologously expressed genes, thereby arriving at “dead-end” local optimum. A “metabolic scaffold” as utilised here can block these dead-end trajectories by coupling a target activity (here rubisco carboxylation) to growth, thereby increasing the likelihood of trajectories that lead to the desired phenotype. Once the phenotype is achieved, removal of the metabolic scaffold should be possible since growth is by definition dependent on the non-native genes. This could serve as a promising strategy for achieving minimally perturbed genotypes in future metabolic engineering attempts.

The approach presented here, of harnessing iterative lab evolution to find a compact set of essential mutations, strikingly shows the extreme malleability of metabolism and evolution capacity to switch trophic modes in laboratory time-scales with very few mutations. It can also serve as a promising pipeline for other genetic engineering attempts. Having a compact and well-defined genotype in a model organism allowed us to reveal metabolic adaptation steps needed for autotrophic growth, and could be used for future exploration of the design principles of the rPP cycle operation and its components(34). The harnessing and rationalisation of the lab evolution process and metabolic adaptation steps elucidated here could facilitate the introduction of the rPP cycle and other carbon fixation cycles into other organisms of interest. Next steps could include integrating metabolic bio-production modules and alternative energy sources. Now that synthetic autotrophy in E. coli can be achieved with only a few defined steps, it expands the playground for more labs to join the quest for sustainable bio-production from CO2.

Supporting information

eLife Assessment

This is an important follow-up study to a previous paper in which the authors reconstituted CO2 metabolism in Escherichia coli (autotrophy). Here, the authors define a set of three mutations that promote autotrophy, highlighting the malleability of E. coli metabolism. The authors make a convincing case that mutations in pgi are loss-of-function mutations that prevent metabolic efflux from the reductive pentose phosphate autocatalytic cycle, but claims about the role of mutations in two other genes - crp and rpoB - are currently incomplete. This research will be particularly interesting to synthetic biologists, systems biologists, and metabolic engineers aiming to develop synthetic autotrophic microorganisms.

Reviewer #1 (Public Review):

The main objective of this study is to achieve the development of a synthetic autotroph using adaptive laboratory evolution. To accomplish this, the authors conducted chemostat cultivation of engineered E. coli strains under xylose-limiting conditions and identified autotrophic growth and the causative mutations. Additionally, the mutational mechanisms underlying these causative mutations were also explored with drill down assays. Overall, the authors demonstrated that only a small number of genetic changes were sufficient (i.e., 3) to construct an autotrophic E. coli when additional heterologous genes were added. While natural autotrophic microorganisms typically exhibit low genetic tractability, numerous studies have focused on constructing synthetic autotrophs using platform microorganisms such as E. coli. Consequently, this research will be of interest to synthetic biologists and systems biologists working on the development of synthetic autotrophic microorganisms. The conclusions of this paper are mostly well supported by appropriate experimental methods and logical reasoning. However, further experimental validation of the mutational mechanisms involving rpoB and crp would enhance readers' understanding and provide clearer insights, despite acknowledgement that these genes impact a broad set of additional genes. Additionally, a similar study, 10.1371/journal.pgen.1001186, where pgi was deleted from the E. coli genome and evolved to reveal an rpoB mutation is relevant to this work and should be placed in the context of the presented findings.

The authors addressed rpoB and crp as one unit and performed validation. They cultivated the mutant strain and wild type in a minimal xylose medium with or without formate, comparing their growth and NADH levels. The authors argued that the increased NADH level in the mutant strain might facilitate autotrophic growth. Although these phenotypes appear to be closely related, their relationship cannot be definitively concluded based on the findings presented in this paper alone. Therefore, one recommendation is to explore investigating transcriptomic changes induced by the rpoB and crp mutations. Otherwise, conducting experimental verification to determine whether the NADH level directly causes autotrophic growth would provide further support for the authors' claim.

Reviewer #2 (Public Review):

Synthetic autotrophy of biotechnologically relevant microorganisms offers exciting chances for CO2 neutral or even CO2 negative production of goods. The authors' lab has recently published an engineered and evolved Escherichia coli strain that can grow on CO2 as its only carbon source. Lab evolution was necessary to achieve growth. Evolved strains displayed tens of mutations, of which likely not all are necessary for the desired phenotype.

In the present paper the authors identify the mutations that are necessary and sufficient to enable autotrophic growth of engineered E. coli. Three mutations were identified, and their phenotypic role in enhancing growth via the introduced Calvin-Benson-Bassham cycle were characterized. It was demonstrated that these mutations allow autotrophic growth of E. coli with the introduced CBB cycle without any further metabolic intervention. Autotrophic growth is demonstrated by 13C labelling with 13C CO2, measured in proteinogenic amino acids. In Figures 2B and S1, the labeling data are shown, with an interval of the "predicted range under 13CO2". Here, the authors should describe how this interval was derived.

The methodology is clearly described and appropriate.

The present results will allow other labs to engineer E. coli and other microorganisms further to assimilate CO2 efficiently into biomass and metabolic products. The importance is evident in the opportunity to employ such strain in CO2 based biotech processes for the production of food and feed protein or chemicals, to reduce atmospheric CO2 levels and the consumption of fossil resources.

Reviewer #3 (Public Review):

The authors previously showed that expressing formate dehydrogenase, rubisco, carbonic anhydrase, and phosphoribulokinase in Escherichia coli, followed by experimental evolution, led to the generation of strains that can metabolise CO2. Using two rounds of experimental evolution, the authors identify mutations in three genes - pgi, rpoB, and crp - that allow cells to metabolise CO2 in their engineered strain background. The authors make a strong case that mutations in pgi are loss-of-function mutations that prevent metabolic efflux from the reductive pentose phosphate autocatalytic cycle. The authors also argue that mutations in crp and rpoB lead to an increase in the NADH/NAD+ ratio, which would increase the concentration of the electron donor for carbon fixation. While this may explain the role of the crp and rpoB mutations, there is good reason to think that the two mutations have independent effects, and that the change in NADH/NAD+ ratio may not be the major reason for their importance in the CO2-metabolising strain.

Specific comments:

1. Deleting pgi rather than using a point mutation would allow the authors to more rigorously test whether loss-off-function mutants are being selected for in their experimental evolution pipeline. The same argument applies to crp.

2. Page 10, lines 10-11, the authors state "Since Crp and RpoB are known to physically interact in the cell (26-28), we address them as one unit, as it is hard to decouple the effect of one from the other". CRP and RpoB are connected, but the authors' description of them is misleading. CRP activates transcription by interacting with RNA polymerase holoenzyme, of which the Beta subunit (encoded by rpoB) is a part. The specific interaction of CRP is with a different RNA polymerase subunit. The functions of CRP and RpoB, while both related to transcription, are otherwise very different. The mutations in crp and rpoB are unlikely to be directly functionally connected. Hence, they should be considered separately.

3. A Beta-galactosidase assay would provide a very simple test of CRP H22N activity. There are also simple in vivo and in vitro assays for transcription activation (two different modes of activation) and DNA-binding. H22 is not near the DNA-binding domain, but may impact overall protein structure.

4. There are many high-resolution structures of both CRP and RpoB (in the context of RNA polymerase). The authors should compare the position of the sites of mutation of these proteins to known functional regions, assuming H22N is not a loss-of-function mutation in crp.

5. RNA-seq would provide a simple assay for the effects of the crp and rpoB mutations. While the precise effect of the rpoB mutation on RNA polymerase function may be hard to discern, the overall impact on gene expression would likely be informative.

Author Response

The following is the authors’ response to the original reviews.

Author response:

Reviewer #1:

The main objective of this study is to achieve the development of a synthetic autotroph using adaptive laboratory evolution. To accomplish this, the authors conducted chemostat cultivation of engineered E. coli strains under xylose-limiting conditions and identified autotrophic growth and the causative mutations. Additionally, the mutational mechanisms underlying these causative mutations were also explored with drill down assays. Overall, the authors demonstrated that only a small number of genetic changes were sufficient (i.e., 3) to construct an autotrophic E. coli when additional heterologous genes were added. While natural autotrophic microorganisms typically exhibit low genetic tractability, numerous studies have focused on constructing synthetic autotrophs using platform microorganisms such as E. coli. Consequently, this research will be of interest to synthetic biologists and systems biologists working on the development of synthetic autotrophic microorganisms. The conclusions of this paper are mostly well supported by appropriate experimental methods and logical reasoning. However, further experimental validation of the mutational mechanisms involving rpoB and crp would enhance readers' understanding and provide clearer insights, despite acknowledgement that these genes impact a broad set of additional genes. Additionally, a similar study, 10.1371/journal.pgen.1001186, where pgi was deleted from the E. coli genome and evolved to reveal an rpoB mutation is relevant to this work and should be placed in the context of the presented findings.

We thank the reviewer for pointing this study out. It is very interesting that a mutation in a similar region in RpoB was observed in a related context of Pgi loss of activity. We have added a reference to this study in our text (Page 11, line 21).

he authors addressed rpoB and crp as one unit and performed validation. They cultivated the mutant strain and wild type in a minimal xylose medium with or without formate, comparing their growth and NADH levels. The authors argued that the increased NADH level in the mutant strain might facilitate autotrophic growth. Although these phenotypes appear to be closely related, their relationship cannot be definitively concluded based on the findings presented in this paper alone. Therefore, one recommendation is to explore investigating transcriptomic changes induced by the rpoB and crp mutations. Otherwise, conducting experimental verification to determine whether the NADH level directly causes autotrophic growth would provide further support for the authors' claim.

We appreciate the valuable comment and agree that the work was lacking such an analysis. Due to various reasons we have opted to use a proteomic approach which we feel fulfills the same purpose as the transcriptomics suggestion. We found interesting evidence in up-regulation of the fdoGH operon (comprising the native formate dehydrogenase O enzyme complex) which could indicate why there is an increase in NADH/NAD+ levels. We also hypothesize that this upregulation might be important more generally by drawing comparisons to natural chemo-autotrophs.

Further experimental work (which we were not able to include in the current study) could help validate this link by deleting fdoGH and observing a loss of phenotype and, on the flip side, directly overexpressing the fdoGH operon and observing an increase in the NADH/NAD+ ratio. Indeed, if this overexpression were to prove sufficient for achieving an autotrophic phenotype without the mutations in the global transcription regulators, it would be a much more transparent design.

We have added a section titled "Proteomic analysis reveals up-regulation of rPP cycle and formate-associated genes alongside down-regulation of catabolic genes" to the Results based on this analysis.

It would be beneficial to provide a more detailed explanation of the genetic background before the evolution stage, specifically regarding the ∆pfk and ∆zwf mutations. Furthermore, it is suggested to include a figure that provides a comprehensive depiction of the reductive pentose phosphate pathway and the bypass pathway. These will help readers grasp the concept of the "metabolic scaffold" as proposed by the authors.

We agree with the reviewer that this could be helpful and we added a reference to the original paper Gleizer et al. 2019 that reported this design and also includes the relevant figure. We feel that the figure should not be added to the current manuscript as we continue to show that this design is not relevant in the context of the three reported mutations and such a figure could distract the attention of the reader from the main takeaways of the current study.

Despite the essentiality of the rpoB mutation (A1245V) to the autotrophic phenotype in the final strain, the inclusion of this mutation in step C1 does not appear to be justified. According to line 37 on page 3, the authors chose to retain the unintended mutation in rpoB based on its essentiality to the phenotype observed in other evolved strains. However, it should be noted that the mutations found in the evolved strain I, II, and III (P552T or D866E) were entirely different from the unintended mutation (A1245V) during genetic engineering. This aspect should be revised to avoid confusion among readers.

Thank you for pointing this issue out, we added a clarification in the text (page 4 line 7) to avoid such confusion. We believe this point is much clearer now.

The rpoB mutation which was shown to be essential in the study is indeed known to be common in ALE experiments in E. coli. Thus, I searched the different rpoB mutations in ALEdb in E. coli and I was able to find a similar mutation in a study where pgi was knocked out and then evolved. https://doi.org/10.1371/journal.pgen.1001186 This study seems very relevant given that pgi was a key mutation in the compact set of this work and the section "Modulation of a metabolic branch-point activity increased the concentration of rPP metabolites" informs that loss of function mutations in pgi were also found. The findings of this study should thus be put in the context of the previous related ALE study. I would recommend a similar analysis of crp mutations from studies in ALEdb to see if there are similar mutations in this gene as well or if this a unique mutation.

We thank the reviewer for bringing this publication to our attention. We have addressed this observation in the main text (page 11 , line 21). We agree that it could have some connection to the pgi mutation yet we would not want to overspeculate about this role, as we also found the exact same mutation (A1245V) as an adaptation to higher temperature in another E. coli study (Tenaillon et al. 2012). We would like to bring forward the fact that the two reported rpoB mutations are always accompanied by another mutation with pleiotropic effects, either in the transcription factor Crp or in another RNA polymerase subunit (e.g RpoC). As such many epistatic effects could occur, one of which we also report here in page 13, line 18. In conclusion, although there could be a connection between the rpoB and pgi mutations, it could be a mere coincidence and the two mutations could exhibit two distinct roles in two distinct phenotypes.

We also would like to thank the reviewer for suggesting a similar analysis for crp and found another mutation at a nearby residue with strong adaptive effects and mentioned it in our main text.

Can the typical number of mutations found in a given ALE experiment be directly compared to those found in this study? It seems like a retrospective analysis of other ALE studies to show how many mutations typically occur in an ALE study and sets which were found to be causal to reproduce the phenotype of interest (through similar reverse engineering in the starting strain) should be presented. Again, the authors cite ALEdb which should provide direct numbers of mutations found in similar ALE studies with E. coli and one could then examine them to find sets of clearly causal mutations which recreate phenotypes of interest. Such an analysis would go a long way in supporting the main finding of "small number" of mutations.

Discussion, page 12, line 42. "This could serve as a promising strategy for achieving minimally perturbed genotypes in future metabolic engineering attempts". There is an entire body of work around growth-coupled production which can be predicted and evolved with a genome-scale metabolic model and ALE. Thus, if this statement is going to be made, relevant studies should be cited and placed in context.

The reviewer raises an important point which could indeed yield an interesting perspective. However, it would be difficult to perform this comparison in practice since many of the studies published on ALEdb have not isolated essential mutations from other mutation incidents nor have they determined the role of each mutation in the reported phenotypes. For example, many ALE trajectories include a hypermutator that greatly increases the number of irrelevant mutations and it is nearly impossible to sieve through them to find an essential set.

Moreover, it is hard to compare the “level of difficulty” of achieving one phenotype over another and therefore feel that even though such an analysis would be insightful, it requires an amount of work which is outside the scope of this study.

Finally, we would like to highlight our approach of using the iterative approach, isolating the relevant consensus mutations and repeating this process until no evolution process is required, we are not aware of prior studies that used this approach.

We now clarified what we mean by "promising strategy" in the discussion in order to avoid any false claims about novelty (page 16 line 32): "Using metabolic growth-coupling as a temporary 'metabolic scaffold' that can be removed, could serve as a promising strategy for achieving minimally perturbed genotypes in future metabolic engineering attempts."

Reviewer #2:

Synthetic autotrophy of biotechnologically relevant microorganisms offers exciting chances for CO2 neutral or even CO2 negative production of goods. The authors' lab has recently published an engineered and evolved Escherichia coli strain that can grow on CO2 as its only carbon source. Lab evolution was necessary to achieve growth. Evolved strains displayed tens of mutations, of which likely not all are necessary for the desired phenotype.

In the present paper the authors identify the mutations that are necessary and sufficient to enable autotrophic growth of engineered E. coli. Three mutations were identified, and their phenotypic role in enhancing growth via the introduced Calvin-Benson-Bassham cycle were characterized. It was demonstrated that these mutations allow autotrophic growth of E. coli with the introduced CBB cycle without any further metabolic intervention. Autotrophic growth is demonstrated by 13C labelling with 13C CO2, measured in proteinogenic amino acids. In Figures 2B and S1, the labeling data are shown, with an interval of the "predicted range under 13CO2".

Here, the authors should describe how this interval was derived.

The methodology is clearly described and appropriate.

The present results will allow other labs to engineer E. coli and other microorganisms further to assimilate CO2 efficiently into biomass and metabolic products. The importance is evident in the opportunity to employ such strain in CO2 based biotech processes for the production of food and feed protein or chemicals, to reduce atmospheric CO2 levels and the consumption of fossil resources.

Please describe in the methodology how the interval of the predicted range of 13C labeling was derived for Figures 2B and S1. Was it calculated by the dilution factor during 4 generations, or did you predict the label incorporation individually with a metabolic model?

The text needs careful editing, some sentences are incomplete and there are frequent inconsistencies in writing metabolites and enzymes.

P2L6: unclear sentence (incomplete?)

P2L19: pastoris with lower case "p"

P2L40: incomplete sentence

P2L42: here, and at many other places, the writing of RuBisCO needs to be aligned. It is an abbreviation and should begin with a capital letter. Most commonly it is written as RuBisCO which I would suggest - please unify throughout the text.

P3L3: formate dehydrogenase ... metabolites and enzymes with lower case letter. And, no hyphen here.

P5L4: delete the : after unintentionally

P6L16: carboxylation of RuBP (it is not CO2 that is carboxylated - if any, CO2 is carboxylating)

P7L25: phosphoglucoisomerase (lower case)

P8L5: in line

P8L9: part of glycolysis/ ...

P10L4: pentose phosphates (lower case, no hyphen).

P10L4: all metabolites lower case

P12L28: incomplete sentence

P18L4: Escherichia coli in italics P18L15: Pseudomonas sp. in italics P18L16: ... promoter and with a strong ...

P20, chapter Metabolomics: put the numbers of 12C and 13C in superscript P23L9: pentose phosphates ; all metabolites in lower case (as above) P23: all 12C and 13C with superscript numbers.

Response to reviewer #2:

We thank the reviewer for their comments, and for pointing out the need to clarify how we derived the predicted range of 13C labeling. We edited the text accordingly, and added the relevant calculation to the methods section (under the “13C Isotopic labeling experiment”). We would like to also thank the reviewer for the required text improvements, which were implemented.

Reviewer #3:

The authors previously showed that expressing formate dehydrogenase, rubisco, carbonic anhydrase, and phosphoribulokinase in Escherichia coli, followed by experimental evolution, led to the generation of strains that can metabolise CO2. Using two rounds of experimental evolution, the authors identify mutations in three genes - pgi, rpoB, and crp - that allow cells to metabolise CO2 in their engineered strain background. The authors make a strong case that mutations in pgi are loss-of-function mutations that prevent metabolic efflux from the reductive pentose phosphate autocatalytic cycle. The authors also argue that mutations in crp and rpoB lead to an increase in the NADH/NAD+ ratio, which would increase the concentration of the electron donor for carbon fixation. While this may explain the role of the crp and rpoB mutations, there is good reason to think that the two mutations have independent effects, and that the change in NADH/NAD+ ratio may not be the major reason for their importance in the CO2-metabolising strain.

We thank the reviewer for their comments and constructive feedback.

We agree that there is probably a broader effect caused by the rpoB and crp mutations, besides the change in the NADH/NAD+ ratio. Hence, we performed a proteomics analysis, comparing the rpoB and crp mutations on a WT background to an autotrophic E.coli, searching for a mutual change in both strains compared to their "ancestors". We found up-regulation of rPP cycle and formate-associated genes, and a down-regulation of catabolic genes. We added a section dedicated to this matter under the title "Proteomic analysis reveals up-regulation of rPP cycle and formate-associated genes alongside down-regulation of catabolic genes".

Specific comments:

1. Deleting pgi rather than using a point mutation would allow the authors to more rigorously test whether loss-off-function mutants are being selected for in their experimental evolution pipeline. The same argument applies to crp.

We appreciate this recommendation and indeed tried to delete pgi, but the genetic manipulation caused a knockout of other genes along with pgi (pepE, rluF, yjbD, lysC) so in the time available to us we cannot confidently determine whether the deletion alone is sufficient and can replace the mutation.

Regarding crp, we do not think there is a reason to believe the mutation is a loss-of-function. In any case, the proteomics-based characterization of the crp mutation is now included in the SI.

1. Page 10, lines 10-11, the authors state "Since Crp and RpoB are known to physically interact in the cell (26-28), we address them as one unit, as it is hard to decouple the effect of one from the other". CRP and RpoB are connected, but the authors' description of them is misleading. CRP activates transcription by interacting with RNA polymerase holoenzyme, of which the Beta subunit (encoded by rpoB) is a part. The specific interaction of CRP is with a different RNA polymerase subunit. The functions of CRP and RpoB, while both related to transcription, are otherwise very different. The mutations in crp and rpoB are unlikely to be directly functionally connected. Hence, they should be considered separately.

Indeed, the fact that the proteins are interacting in the cell does not necessarily mean that the mutations are functionally connected. We therefore added as further justification in the new section:

"As far as we know, the mutations in the Crp and RpoB genes affect the binding of the RNA polymerase complex to DNA and/or its transcription rates. Depending on the transcribed gene target, the effect of the two mutations might be additive, antagonistic, or synergistic. Since each one of these mutations individually (in combination with the pgi mutation) is not sufficient to achieve autotrophic growth, it is reasonable to assume that only the target genes whose levels of expression change significantly in the double-mutant are the ones relevant for the autotrophic phenotype”.

In our proteomics analysis we considered each mutation separately. We found that in some cases the two mutations together have an additive effect, but in other cases we found that the two mutations together affect differently on the proteome, compared to the effect of each mutation alone. Since both mutations are essential to the phenotype, we decided to go with the approach of addressing the two mutations as one unit for the physiological and metabolic experiments.

1. A Beta-galactosidase assay would provide a very simple test of CRP H22N activity. There are also simple in vivo and in vitro assays for transcription activation (two different modes of activation) and DNA-binding. H22 is not near the DNA-binding domain, but may impact overall protein structure.

The mutation is located in “Activating Region 2”, interacting with RNA polymerase. We tried an in-vivo assay to determine the CRP H22N activity and got inconclusive results, we believe the proteomics analysis serves as a good method for understanding the global effect of the mutation.

1. There are many high-resolution structures of both CRP and RpoB (in the context of RNA polymerase). The authors should compare the position of the sites of mutation of these proteins to known functional regions, assuming H22N is not a loss-of-function mutation in crp.

We added a supplementary figure regarding the structural location of the two mutations, where it is demonstrated that crp H22N is located in a region interacting with the RNA polymerase and rpoB A1245V is located in proximity to regions interacting with the DNA.

1. RNA-seq would provide a simple assay for the effects of the crp and rpoB mutations. While the precise effect of the rpoB mutation on RNA polymerase function may be hard to discern, the overall impact on gene expression would likely be informative.

Indeed we agree that an omics approach to infer the global effect of these mutations is beneficial, we opted to use a proteomics approach and think it serves the purpose of clarifying the final, down-stream, effect on the cell.

1. Page 2, lines 40-45, the authors should more clearly explain that the deletion of pfkA, pfkB and zwf was part of the experimental evolution strategy in their earlier work (Gleizer et al., 2019), and not a new strategy in the current study.

We thank you for pointing this out, and edited the text accordingly.

1. Page 3, line 27. Why did the authors compare the newly acquired mutants to only two mutants from the earlier work, not all 6?

The 6 clones that were isolated in Gleizer et al., had 2 distinct mutation profiles. During the isolation process the lineage split into two groups. Three out of the 6 clones (clones 1,2,6) came from the same ancestor, and the other three (clones 3,4,5) came from another ancestor. Hence, these two groups shared almost all of their mutations (see Venn diagram). We decided to use for our comparison the representative with the highest number of mutations from each group (clones 5 and 6).

Version 2

Abstract

Abstract

Synthetic autotrophy is a promising avenue to sustainable bioproduction from CO2. Here, we use iterative laboratory evolution to generate several distinct autotrophic strains. Utilising this genetic diversity, we identify that just three mutations are sufficient for E. coli to grow autotrophically, when introduced alongside non-native energy (formate dehydrogenase) and carbon-fixing (RuBisCO, phosphoribulokinase, carbonic anhydrase) modules. The mutated genes are involved in glycolysis (pgi), central-carbon regulation (crp), and RNA transcription (rpoB). The pgi mutation reduces the enzyme’s activity, thereby stabilising the carbon-fixing cycle by capping a major branching flux. For the other two mutations, we observe down-regulation of several metabolic pathways and increased expression of native genes associated with the carbon-fixing module (rpiB) and the energy module (fdoGH), as well as an increased ratio of NADH/NAD+ - the cycle’s electron-donor. This study demonstrates the malleability of metabolism and its capacity to switch trophic modes using only a small number of genetic changes and could facilitate transforming other heterotrophic organisms into autotrophs.

Introduction

Anthropogenic interference and intensive burning of fossil fuels releases large amounts of sequestered carbon in the form of CO2, a main driver of climate change (Olofsson and Hickler 2008; Friedlingstein et al. 2022). Direct air carbon capture, by any means, is urgently needed to help reduce emissions. Here, we focus on biology-based carbon fixation as a potential avenue for CO2 capture, which has been shown to be a potentially efficient production platform (Leger et al. 2021; Van Peteghem et al. 2022; Schwander et al. 2016). Specifically, we suggest using microbes for CO2 valorization and produce a diverse repertoire of products including food, fuel, bio-plastics or other commodities (Claassens et al. 2019; Gleizer et al. 2020; Li et al. 2012; Nielsen and Keasling 2016). Natural autotrophs already metabolise CO2 efficiently, yet most are challenging to cultivate and manipulate genetically compared to heterotrophic model organisms like E. coli. Here we explore an emerging alternative approach – converting heterotrophic model organisms into tractable autotrophs for bio-production (Gleizer et al. 2020; Baumschabl et al. 2022; Gassler et al. 2020; Gleizer et al. 2019). In addition to promising engineering applications, studying the transition from heterotrophic to autotrophic metabolism can uncover fundamental principles about the structure and regulation of carbon fixation metabolism (Gleizer et al. 2019; Gassler et al. 2020; Antonovsky et al. 2016; Barenholz et al. 2017; Flamholz et al. 2020; Herz et al. 2017).

Recently, with the increasing interest in platforms for sustainable production, synthetic biologists have been able to successfully manipulate different organisms to enable major metabolic transitions (Gassler et al. 2020; Gleizer et al. 2019; Kim et al. 2020; Chen et al. 2020; Keller et al. 2022; Bang et al. 2020; Satanowski et al. 2020; Yishai et al. 2018). These efforts resulted in the successful introduction of non-native C1 utilisation pathways, and even included successful transitions from heterotrophic into completely autotrophic metabolisms in bacteria (Gleizer et al. 2019) and yeast (Gassler et al. 2020). These previous studies introduced the reductive pentose phosphate (rPP) cycle, the most prominent carbon fixation pathway in nature (also known as the Calvin-Benson cycle), into Escherichia coli (E. coli) and Komagataella phaffii (previously Pichia pastoris). Indeed, we showed that E. coli was able to utilise the rPP cycle for the synthesis of all biomass carbon - thus converting it from a heterotroph to an autotroph (Gleizer et al. 2019). However, this transition required continuous culture in selective conditions for several months (adaptive lab evolution) and resulted in many uninterpretable mutations, exposing a large knowledge gap.

Using adaptive lab evolution presents a challenge as it is hard to isolate the changes necessary for a given phenotype from the overall set of accumulated mutations. Here we aimed to characterise the landscape of mutations required for autotrophic growth. Studying mutations that arise in multiple lineages can help us better understand the biological basis for the heterotroph-to-autotroph transition. To achieve this goal, we created a pipeline that enabled us to distinguish mutations essential for autotrophy from unspecific low-frequency mutations, i.e. ones that are generally promoted by the physiological conditions in evolution experiments. The final outcome of this pipeline is a compact essential set of changes permitting autotrophy in E. coli. We then proceeded with further analyses to shed light on the mechanism behind this compact set of essential mutations. We show the phenotype included a change in the activity of a central carbon metabolism enzyme and intracellular changes of the non-native carbon fixation cycle’s metabolite and co-factors pools. We conclude by speculating about the role of these metabolic changes.

Results

A compact set of mutations that enable autotrophic growth was identified using rational design, iterative lab evolution and genetic engineering

Adaptive lab evolution is an effective tool that enables integration of non-native metabolic pathways. In order to harness the power of adaptive lab evolution, a selection system must be implemented to direct the evolution towards the desired function. In our case, the selection for the activity of the non-native rPP cycle. To ensure that some level of carbon fixation by RuBisCO becomes essential for growth, we knocked out phosphofructokinase (pfkA and pfkB), and 6-phosphate-1-dehydrogenase (zwf) which creates a stoichiometric imbalance and growth arrest when growing on five carbon sugars such as xylose (Gleizer et al. 2019). This imbalance could be rescued by a metabolic bypass - the phosphorylation of ribulose-5-P by Prk and the subsequent carboxylation by RuBisCO, thereby coupling growth to the heterologously expressed rPP enzymes. Critically, this setup allowed us to evolve E. coli, a natural heterotroph, into an autotroph, using a non-native rPP cycle to utilise CO2 as a sole carbon-source. This system also requires formate dehydrogenase to oxidise formate for energy (Gleizer et al. 2019).

Although adaptive lab evolution is a powerful tool, it also presents a challenge - once a new phenotype has been achieved, it is hard to separate the essential changes required for its appearance from the overall set of accumulated mutations. In many cases, mutations occur on the background of highly adaptive ones, leading to a spread of neutral “hitchhiker” mutations in the population. In other cases, mutations that appeared at the beginning of the evolution are no longer needed for the final phenotype. Our goal was to find a set of mutations that, if introduced into a heterotrophic wild type E. coli, would result in the successful integration of the non-native carbon fixation and energy modules. This would enable autotrophic growth based solely on genetic engineering without the need for further evolutionary adaptation and potentially uncovering the mechanism behind this metabolic transition.

To isolate the essential genetic changes, we used a workflow that included three main stages as shown in Fig. 1. A) rational design-introduction of required heterologous genes and metabolic knockouts to enforce RuBisCO-dependent growth; B) adaptive lab evolution-revealing mutation candidates by growing the designed strain in autotrophic-selecting conditions over many generations until an autotrophic phenotype is obtained; C) evolution-inspired genetic engineering-introducing the most promising mutations revealed in stage B into the designed strain and testing it for autotrophic growth. Stages B and C were repeated until no further evolution was required, i.e. the designed strain in stage C had an autotrophic phenotype.

Autotrophic phenotype achieved by introducing 3 mutations on top of a rationally designed ancestor.

(A) We rationally designed the “wild-type” E. coli background strain (BW25113, depicted as a black bacterium) by introducing 4 enzymes: RuBisCO (cbbM), phosphoribulokinase (prkA), carbonic anhydrase (CA) and formate dehydrogenase (fdh), and 3 genomic knockouts: glucose-6-phosphate dehydrogenase (zwf) and phosphofructokinase A&B (pfkAB). We denote the resulting strain as “ancestor” (brown bacterium). (B) We tested the resulting strain for autotrophic growth and, since it didn’t grow, we used it for adaptive lab evolution in xylose-limited chemostats - i.e., conditions that select for higher carbon fixation flux. Altogether, we were able to isolate three evolved clones (two distinct strains from one chemostat experiment, Evolved I & II, and another strain from a second experiment, Evolved III, step B1). (C) Out of four consensus mutations that we identified, two - rpoB and pgi - were incorporated into the ancestor, giving “ancestor 2.0” (light blue bacterium, step C1). Once more, we tested for autotrophic growth and were unsuccessful. Therefore, we initiated another round of adaptive lab evolution experiments using ancestor 2.0 growing in two xylose limited chemostats and isolated two evolved clones (one from each chemostat, step B2). The two clones shared a single consensus mutation, in crp. We thus created a new strain from ancestor 2.0 by introducing the crp mutation into it (blue bacterium, step C2). This strain could grow autotrophically and thus we achieved a compact autotrophic strain.

To find the most promising mutations in stage B, we applied at least one of two criteria: mutations in genes that occur repeatedly in different autotrophic lineages (but are uncommon in other adaptive lab evolution experiments) or mutations that when reverted, result in a loss of phenotype. We discovered these “consensus” mutations by sequencing isolated autotrophic clones, obtained from repeated adaptive lab evolution experiments-similar to the conditions we used in Gleizer et al. 2019 (Fig. 1; steps B1, B2).

In practice, we inoculated the strain denoted as Ancestor into a chemostat with limited xylose and excess formate. After ≈3 months (≈60 chemostat generations) we isolated from the chemostat a clone able to grow under autotrophic conditions (methods), and compared its mutation set to those of two previously evolved autotrophic clones (Gleizer et al. 2019). The intersection of the three clones contained 4 genes that were mutated in all autotrophic isolated clones: the central carbon metabolism enzyme phosphoglucoisomerase (pgi), the beta subunit of the RNA polymerase (rpoB), poly A polymerase (pcnB) and the MalT DNA-binding transcriptional activator (malT) (Fig. 1, step B1). rpoB, pcnB and malT are genes that are commonly mutated in adaptive lab evolution experiments (Phaneuf, Gosting, and Palsson 2019), suggesting that they have little relevance for autotrophic growth. Therefore, we chose to focus on the pgi mutation.

We introduced the H386Y pgi mutation into the rationally designed ancestor (Fig. 1, step C1), and the engineered strain was tested for growth in autotrophic conditions (methods). No growth was observed in those conditions, which meant that additional mutations were necessary in order to achieve autotrophy.

Genomic sequencing revealed that during the genetic manipulation for introducing the pgi mutation, another mutation in rpoB (A1245V) appeared in the genome. In parallel, reverting a variety of RNA polymerase mutations in other evolved strains back to their wild-type allele, showed that mutations in RNA polymerase are in fact essential to the phenotype. Therefore, despite the fact that it was unintentional, and different from the other rpoB mutations observed in the ALE experiments, we left the mutation in the genome and proceeded with this strain.

This strain, mutated in both pgi and rpoB (denoted “ancestor 2.0”) was then used as the starting point for another round of evolution (Fig. 1, step B2). Since it had already contained two mutations that were likely to be relevant for the desired phenotype, we expected to observe evolved autotrophic strains after fewer generations, compared to the first round, and to find fewer mutations in them. Indeed, on two separate experiments, it took roughly one month until autotrophic growth was observed, as opposed to more than 90 days in previous attempts. Even though these strains (Fig. 1, “Evolved 4” and “Evolved 5”) both had multiple mutations, only one mutation was shared between these independent evolution experiments - a non-synonymous mutation in the cAMP receptor protein, crp (H22N). When we introduced this mutation to the “ancestor 2.0” background in the evolution-inspired genetic engineering stage (Fig. 1, step C2), the strain was immediately able to grow under autotrophic conditions.

Since the RNA polymerase mutation was introduced into this strain (rpoB A1245V) as a byproduct of genetic engineering and never observed in the adaptive evolution experiments, we needed to verify its essentiality to the autotrophic phenotype. We found that a reversion of rpoB A1245V back to the wild-type allele causes a loss of the autotrophic phenotype. We therefore could conclude that three mutations (pgi*, crp* and rpoB*) are sufficient to facilitate the autotrophic phenotype on top of the heterotrophic ancestor background. We denote the final engineered autotrophic E. coli strain, the “compact autotrophic E. coli”.

We verified the genotype using whole genome sequencing (methods). The results included the rationally designed knockouts (ΔpfkA, ΔpfkB, Δzwf), heterologous plasmids (energy and carbon-fixing modules) and the three introduced mutations (pgi*, crp* and rpoB*). During the genetic engineering process, two additional mutations occurred unintentionally in the genes uhpT and yejG. uhpT is a hexose transporter and is unlikely to have any effect on the autotrophic phenotype, especially because the mutation was an early stop codon (Q7*) and likely a loss of function. As the function of yejG is unknown, we wanted to ensure that it is not essential to the autotrophic phenotype. Therefore, we reverted the mutation back to its wild-type allele and found that indeed the cells were still autotrophic.

Thus, we constructed a new autotrophic strain with three essential mutations (pgi, crp & rpoB), which we denoted the compact strain. When comparing this compact strain to a previously characterised autotrophic strain (Gleizer et al. 2019), it was nearly identical in terms of growth rate, lag time and yield (Fig. 2A). Using 13C labelling, we verified that all the biomass carbon stems from CO2 (Fig. 2B). We termed the three introduced mutations (pgi, crp & rpoB) together with the carbon fixation machinery (cbbM, prkA & CA) and the energy module (fdh) the “autotrophic enabling gene set”.

Validation and characterisation of the autotrophic phenotype.

(A) left: growth curve of an isolated evolved clone vs the engineered compact strain in liquid M9 minimal media supplemented with 45mM sodium formate and sparged with a gas mixture of 10% CO2, 5% oxygen, and 85% nitrogen. right: calculated growth rate using linear regression of the engineered compact strain (light blue) and evolved strain (orange), the calculated doubling time with the given conditions is about 24h (μ≈0.03h-1). Growth was carried out in triplicates (n=3) in a Spark plate reader with gas control, dark grey area corresponds to the standard deviation of the mean (150 μl culture + 50 μl mineral oil, to prevent evaporation). (B) The fractional contribution of 13CO2 to various protein-bound amino acids of the compact autotrophic strain after ≈4 doublings on 13CO2 and 13C labelled formate (light blue, mean of n=3). This reached the expected 13C labelling fraction (see methods) of the biosynthesized amino acids. When grown in naturally labelled CO2 and 13C labelled formate (dark blue, n=1, technical triplicate) the 13C content dropped to close to natural abundance. Experiments with 13CO2 as the substrate were carried out in air-tight (i.e., sealed) growth vessels.

Introduction of autotrophy-enabling gene set into wild-type background does not require bypass-prevention for growth

As described above, in order to select for autotrophic growth, the native E. coli was metabolically re-wired by 3 auxiliary genomic knockouts (ΔpfkA, ΔpfkB, Δzwf). These knockouts created a dependency on carboxylation by RuBisCO for growth even when consuming pentose sugars which were supplemented during the chemostat evolution (see methods, Antonovsky et al. 2016; Gleizer et al. 2019). This dependency was expected to direct the evolution towards increased usage of the synthetic rPP cycle and, eventually, to autotrophic growth.

Once autotrophic growth was achieved, and the enabling gene set was found, we wanted to test if these auxiliary knockouts (pfkA pfkB and zwf) were still required for the phenotype. We introduced the set of autotrophy-enabling genes into a wild-type (BW25113) E. coli background (methods, Fig. 3). After doing so, we discovered an unintentional genomic mutation, in the ribosomal 16S subunit (rrsA) that was introduced during the genomic manipulation process. The engineered E. coli strain was able to grow autotrophically, although its final OD was half and its growth rate was less than half of that of the rationally designed compact strain still harbouring the auxiliary knockouts (Fig. S2). This indicates that the “autotrophic enabling gene set” is sufficient to achieve the trophic mode change and that the knockouts are helpful but not essential for the final autotrophic growth. As discussed below, this demonstrates that, in some cases, metabolic rewiring could be viewed as a temporary “metabolic scaffold” which can be removed from the genotype once the desired phenotype is achieved.

Auxiliary knockouts are not required for final phenotype.

We transformed a wild-type BW25113 E. coli strain (black bacteria) with the carboxylating and energy module plasmids (grey circles with coloured gene annotations), and inserted 3 auxiliary genomic knockouts (red octagon) to rewire metabolism toward carboxylating dependency. This strain was used for iterative evolution experiments in order to generate diverse autotrophic strains and reveal mutations candidates for rational design, i.e. consensus mutations. The identified autotrophic enabling mutations (thin grey circle) were then introduced in a wild type strain expressing the heterologous plasmids but without the auxiliary knockouts, the final strain was able to grow in autotrophic conditions. Dashed lines represent gene/mutation introduction.

We thus established a new engineered autotrophic E. coli strain containing a set of three essential and sufficient mutations, and showed that the bypass-prevention knockouts introduced before the ALE are themselves not necessary for the phenotype. Due to the small required number of genetic modifications, this strain provides an opportunity to derive guidelines for future engineering efforts that use the rPP cycle. Having this goal in mind, we focused our attention on identifying the relevant phenotypes of each of these mutations and explaining their adaptive advantage. Because the mutations were essential for the autotrophic growth, we could not use autotrophic conditions to compare them to a strain with wild-type alleles. Therefore, we chose the most suitable combinations of strains and conditions that could isolate the effect of the pertinent mutation.

Modulation of a metabolic branch-point activity increased the concentration of rPP metabolites

Metabolic pathways that regenerate and synthesise more of their own metabolites are referred to as autocatalytic cycles. Within the autotrophic E. coli the rPP cycle is autocatalytic. Due to the inherent positive feedback mechanism, autocatalytic cycles tend to be unstable, and therefore the fluxes of entry and exit points (bifurcation/branch points) need to be balanced (Barenholz et al. 2017). Such tuning is needed since any disruption of the balance between cycling and branching fluxes could result in the depletion, or alternatively toxic accumulation of intermediate metabolites, and cycle arrest.

We previously predicted that mutations in branch points will be needed to stabilise the steady state flux within the rPP cycle (Barenholz et al. 2017). In line with this prediction we find that the mutation in phosphoglucoisomerase (Pgi), a key branch point in the rPP cycle, follows this design principle. Pgi consumes fructose-6-phosphate (F6P), one of the cycle intermediates, converting it into glucose-6-phosphate (G6P), a precursor for cell membrane biosynthesis. Thus, Pgi regulates flux out of the autocatalytic cycle (Fig. 4A).

Pgi mutation is required to partition flux towards regeneration of autocatalytic cycle substrate.

(A) Metabolic scheme of rPP cycle components when growing on xylose. (B) In vitro spectrophotometric coupled assay determined that the rate of the isomerization of fructose-6-phosphate (F6P) to glucose-6-phosphate (G6P) is ≈100-fold lower for purified PgiH386Y (≈1.5 [s-1]) compared to Pgiwild-type (≈130 [s-1]), n=3. Noise level was determined by negative control samples containing a non-Pgi enzyme. (C) Left: The wild type pgi, competes for its substrate F6P with the autocatalytic cycle, resulting in low F6P pool and low regeneration rate of ribulose-5-phosphate (Ru5P), which means that the RuBisCO-dependent pathway requires constant xylose supply. Right: The mutated Pgi (green) has reduced activity, which increases the ratio between the flux in the cycle and efflux which is needed for a stable regenerative flux towards Ru5P (via the pentose-phosphate-pathway), thus enabling an autotrophic cycle. (D) Left: Measured relative intracellular ratio of F6P and G6P of the ancestor background harbouring a PgiH386Y mutation compared to the ancestor with a wild-type Pgi, both grown in RuBisCO dependent conditions, n=6 cultures p-value < 0.05. Right: Relative intracellular abundance of pentose-phosphate-pathway metabolites - sedoheptulose-7P (S7P) and total pool of pentose phosphates (ribulose-5P, ribose-5P, ribose-1P, and xylulose-5P - denoted P5P) in a PgiH386Y strain versus the Pgiwild-type strain, both growing in a RuBisCO-dependent manner, n=3 cultures.

In the different adaptive lab evolution experiments, we found three distinct mutations in the pgi gene. The first, H386Y, is a non-synonymous mutation occurring in one of the catalytic residues in the active site, and is part of the autotrophic enabling gene set. The second was a complete knockout, a 22KB chromosomal deletion, including also 16 other genes. The third was an early stop codon E72*. These observations led us to suspect the H386Y mutation in pgi decreases or even completely eliminates the activity of the enzyme.

We tested the kinetic rates of the isomerization of F6P to G6P of two purified Pgi enzymes, wild type Pgi and Pgi H386Y, using a spectroscopic coupled assay, where G6P was coupled to NADP+ reduction (Widjaja et al. 1998) (Fig. 4B; see also methods). The measured kcat of the wild type Pgi was ≈130 [s-1] (±26) which is in line with previous computational predictions of a kcat of ≈200 [s-1](Davidi et al. 2016). The mutated Pgi showed weak activity with a kcat of ≈1.5 [s-1](±1). Therefore the pgi mutation is in line with the original expectation that branching point regulation is necessary, and the nature of the observed mutations in pgi supports the notion that their role is to reduce the efflux from the non-native autocatalytic cycle.

Pgi is part of glycolysis/gluconeogenesis, where significant flux is required in wild type E. coli. Accordingly, wild type Pgi catalyses a reaction that is causing a strong efflux of F6P from the cycle, diminishing the regeneration of ribulose-1,5-bisphosphate (RuBP) which is usually not needed in the native metabolism but is essential for the rPP autocatalytic activity. In the mutated version of Pgi the efflux capacity is diminished, which can stabilise the cycle (Fig. 4C).

Following the same logic, we can expect an increase in the F6P substrate metabolite pool due to reduced consumption by Pgi. Therefore, we measured intracellular sugar-phosphates in the pgi mutant strain and a strain with a wild-type pgi allele by liquid chromatography-tandem mass spectrometry (LC-MS/MS). For this comparison, we used the RuBisCO-dependent ancestor strain as the genetic background. We found that the ratio of F6P to G6P was about 3 times higher in the pgi mutant strain relative to the wild-type (Fig. 4D). Furthermore, the pgi mutant had higher levels of metabolites within the rPP cycle (Fig. 4D), confirming the stabilising function of the mutation.

These results lead us to conclude that the recurring mutations in the pgi gene are mediating the integration of the non-native carbon fixation genes into a stable autocatalytic cycle. This is achieved by reducing the ratio of the efflux from F6P to G6P relative to the flux in the cycle, thereby redirecting it towards the rPP cycle and the regeneration of RuBP.

Proteomic analysis reveals up-regulation of rPP cycle and formate-associated genes alongside down-regulation of catabolic genes

The other two mutations in the compact set - crp (H22N) and rpoB (A1245V), are global effectors in the cell. Crp is one of the most general transcription factors in the cell and has a role in many processes from metabolism to osmoregulation and biofilm formation (Landis, Xu, and Johnson 1999; Jackson, Simecka, and Romeo 2002; Ammar, Wang, and Rao 2018). The position H22 in Crp is part of “activation region 2” of the protein, an interaction region with RNA polymerase (Fig. S8). This position was shown to have an effect on promoter specificity (Niu et al 1996) and was previously reported to have a role in the integration of the rPP cycle in E. coli (Herz et al. 2017).

Similarly, mutations in rpoB, a subunit of RNA polymerase, have been shown to have a large effect on the cell transcriptome (Conrad et al. 2010; Utrilla et al. 2016). The observed A1245V mutation in rpoB is situated within the core of RNA polymerase, proximal to its interaction with the DNA (figure S8). Furthermore, this genetic variation was previously documented in the adaptation of the E. coli B REL1206 strain to high temperatures (Tenaillon et al. 2012) and is closely positioned to another previously reported rpoB mutation (K1242Q) that appeared in adaptation to loss of Pgi activity (Charusanti et al. 2010).

Depending on the transcribed gene target, the effect of the two mutations might be additive, antagonistic, or synergistic. Since each one of these mutations individually (in combination with the pgi mutation) is not sufficient to achieve autotrophic growth, it is reasonable to assume that only the target genes whose levels of expression change significantly in the double-mutant are the ones relevant for the autotrophic phenotype. Given the accumulated knowledge described above, these mutations are likely to have a broad effect on the cell. We chose to further elucidate their effect using a proteomics approach. As such, we conducted two experiments in an attempt to find changes in expression that are relevant for the adaptation to the new trophic mode. The first experiment compared the proteome of an evolved clone to its ancestor. The second experiment similarly compared an engineered BW25113 (BW) strain, containing the two regulatory mutations from the compact set (i.e., crp H22N and rpoB A1245V), to a wild type BW strain.

Since our aim was to observe only direct effects induced by the genetic changes, rather than by the environment, we measured the proteomes of the strains under identical conditions and dilution rates. Ideally, the comparison should be done in conditions that require autotrophic growth (since some effects might depend on the lack of an organic carbon source), however, all strains except the evolved clone cannot grow in such conditions. Therefore, we decided on growing the cells in the conditions present at the beginning of the lab evolution experiment, namely a xylose-limited chemostat. The dilution rate was set to ≈0.03 h-1 leading to a doubling time of ≈24h (methods).

In order to identify the genes with a large and significant fold change (fold change >4, p-value <0.05) we generated a volcano plot for each such comparison (Fig. S6). We then proceeded by comparing the output of the first experiment (evolved divided by ancestor, Fig. S6A) to that of the double mutant divided by the base BW strain (Fig. S6B), since both mutations are essential for the phenotype and are known to affect expression. We focused on genes with common trends of expression level in both sets (Fig. 5).

Proteomics analyses of fold change in evolved/ancestor expression levels versus double mutant/WT.

Comparison of the log2 fold change in the evolved autotrophic strain divided by the ancestor expression versus BW rpoB A1245V & crp H22N, divided by the WT. The size of dots represents significance, only genes with >4 fold-change who showed significance (<0.05 p-value) were annotated.

The proteomics analysis yielded few genes that had an upregulation of >4-fold in both the evolved versus ancestor and the double mutant versus BW base strain experiments. Two of these up-regulated genes, fdoG and fdoH, are subunits of the native formate dehydrogenase O enzyme complex (FDH-O) and are therefore related to the energy module. These genes were upregulated in both the evolved strain and in the strain with the crp H22N and rpoB A1245V double mutation (≈5-fold increase in both cases). The FDH-O enzyme complex takes part in the respiratory chain formed by formate dehydrogenase and dimethyl sulfoxide (DMSO) reductase. Here, the transfer of electrons from formate to DMSO (via a quinone electron carrier) is coupled to the generation of a proton-motive force across the cytoplasmic membrane for energy formation.

Another observed upregulated protein was ribose-phosphate-isomerase B (rpiB), which was up-regulated in the evolved strain (>40-fold increase) and in the double mutant ≈5-fold increase (Fig. 5). rpiB is an inducible isoenzyme of the constitutively expressed ribose-phosphate-isomerase A, a core enzyme in the rPP cycle. This might serve as evidence for a higher flux requirement through the pentose-phosphate-pathway, supported by expressing a second isoenzyme. Other explanations may include some form of regulation, perhaps allosteric, inhibiting RpiA activity, making the strain dependent on RpiB activity, or non-specific enzymatic activity of RpiB which is required by the cell. The fourth up-regulated gene - ftsI, a peptidoglycan DD-transpeptidase - does not have a clear explanation at this time.

At the other end of the expression spectrum, several pathways were down-regulated in both the evolved and in the BW double mutant strain (Fig. 5). Down-regulation of the starvation response, metabolic and PTS component genes seems to be at the heart of the adaptation to autotrophic growth, with several operons experiencing >4-fold decrease in expression. Most notably the gat, ara and man operons responsible for galactitol, arabinose and mannose degradation, respectively. Again, this is evident in both the evolved and the double mutant strains (Fig. 5). Other highly down-regulated metabolic pathways include amino acid biosynthesis, specifically valine and leucine, moreover we observed a down regulation in rspA & rspB, which encode for starvation sensing proteins. We suggest these results mean that many pathways usually up-regulated in E. coli during starvation may be redundant or even counter-productive in an autotrophic context, and decreasing their expression could result in increased fitness.

We additionally tested whether separating between the two regulatory mutations (crp and rpoB) might provide a clearer picture, e.g. if different up- or down-regulated genes are targeted by only one of these mutations and not the other. Therefore, we used single-mutant BW strains and compared them to the same BW wild-type as before. Essentially, the results did not change our conclusions and although in most cases the effect of the two mutations was cumulative, we did find at least one case where they were antagonistic (Fig. S7). For example, fdoG and fdoH showed a ≈4-fold decrease compared to wild type when the crp mutation was introduced, but (as stated earlier) had a ≈5-fold abundance increase in the double mutant, suggesting an epistatic effect between the rpoB and crp mutations (Fig. S6B & Fig. S7B). All the comparisons based on the single and double mutant strains can be found in the supplementary.

Mutations in regulatory genes lead to increased availability of the carbon fixation cycle electron donor NADH

Along with the effect on expression, we suspected the two transcription-associated mutations (crp & rpoB) to also have a metabolic and physiological effect. Since Crp and RpoB are both essential to the autotrophic phenotype and are known to indirectly interact in the cell (Savery, Rhodius, and Busby 1996; Niu et al. 1996; Frendorf et al. 2019), we address them as one unit. The proteomics analysis, where the strain with both mutations seem to have a mutual effect that is in some cases different than in the individual mutants, supported this decision. Mutations in crp and rpoB are known to affect growth at different conditions such as types of media or carbon sources (Utrilla et al. 2016; Conrad et al. 2010; Cheng et al. 2014; Frendorf et al. 2019). Therefore we tested whether this was also the case for the conditions under which the cells evolved in the chemostat.

We performed a growth experiment of the double mutant (crp H22N, rpoB A1245V) expressing fdh in a minimal medium containing xylose and formate. Under such conditions, the strain harbouring both mutations and fdh exhibited >25% increase in the final OD compared to the wild type expressing fdh (Fig. 6A&B). At the same time they also showed a longer (>10 hours) lag time. Notably, when grown solely on xylose, in the absence of formate, the increase in yield was not observed, and the extended lag was less pronounced (Fig. 6A).

mutations in rpoB and crp increase yield and intracellular NADH/NAD+ levels in fdh-expressing E. coli in the presence of formate.

Growth experiment of BW25113 wild-type (grey) compared to a crp H22N rpoB A1245V mutant (orange). Both strains express fdh. (A) strains were grown in 1g/L xylose and 40mM formate (solid line) or in the absence of formate (dashed line). The experiment was executed in a 96 well plate in 10% CO2 atmosphere, n=6 repeats. Lines represent the mean, light-grey background represents the standard deviation. (B) Maximal OD600 of the wild-type (grey) and mutant (orange) strains grown in 1g/L xylose and 40mM formate or in the absence of formate. Bar heights represent averages (n=6) of the median of the top 10 OD measurements of each replicate. Error bars represent standard error. (C) Intracellular NADH/NAD+ ratio of the wild-type (grey) was compared to the mutant (orange). The strains were grown in 2g/L xylose and 30mM formate, or in the absence of formate. The y-values are NADH/NAD+ ratios as fold-changes relative to wild-type without formate. Boxes represent 25-75 percentile ranges and dark lines represent median values. All data points are depicted without removing outliers.

Since NADH is the electron donor for carbon fixation, and a product of formate consumption via Fdh, we used metabolomics to measure the ratio of NADH/NAD+ in these strains in similar conditions (methods). All of the measured ratios were normalised to the wild-type (with fdh) growing in the absence of formate. The addition of 30mM formate, which can serve as an electron donor via oxidation to CO2 by Fdh, increased NADH/NAD+ ratio by >20-fold in the double mutant, significantly higher than the ≈10-fold increase in the wild-type strain (p-value <0.05). No significant difference was observed between the strains in the absence of formate (Fig. 6C).

NADH could affect growth in several ways, as discussed below. The specific reason for the increase in final OD or lag time remains inconclusive at this point but we hypothesise that it is related to the increased expression of Fdo observed in the proteomic analysis. We also note that as the proteomics results suggest - the mutations in global transcriptional machinery like Crp & RpoB could have many other effects on the autotrophic phenotype (Haverkorn van Rijsewijk et al. 2011) which await future exploration.

Discussion

In this study, we demonstrated how iterative evolution and engineering were able to narrow down the genomic mutations required for autotrophy in E. coli from tens of mutations to only three mutated genes. The three mutated genes - pgi, crp and rpoB - serve as a compact set, which when introduced into a wild-type background and supplemented with the carbon-fixing and energy modules, achieve an autotrophic phenotype. By metabolic, proteomic and physiological characterisation we were able to distil some of the effects of these mutations and understand their role in facilitating the autotrophic phenotype. We suggest that the three main effects achieved by the autotrophy-enabling mutations are: (i) integrating the non-native carbon fixation cycle by enhancing flux within the cycle over outgoing flux; (ii) facilitating the integration of the energy module via decoupling of formate harvesting into energy and redox, thereby increasing the NADH/NAD+ ratio, which could have both thermodynamic and kinetic benefits (iii) and freeing protein resources by decreasing unneeded expression.

The mutation in pgi affects the activity of its gene product phosphoglucoisomerase - a metabolic enzyme that catalyses a reaction that branches out of the non-native rPP cycle. The enzyme converts F6P (a cycle intermediate) to G6P (a precursor required for membrane biosynthesis). In a previous study, this mutation was found to be essential for the integration of the rPP autocatalytic cycle and was part of a minimal set for hemi-autotrophic growth (Gleizer et al. 2019; Gassler et al. 2020; Antonovsky et al. 2016; Barenholz et al. 2017; Flamholz et al. 2020; Herz et al. 2017). As our results suggest, the mutated enzyme shows a significant reduction in activity, which decreases the efflux from the rPP cycle to biosynthesis. We hypothesised that its role is to generate a new, higher, steady-state concentration for the rPP cycle metabolites, thus stabilising the cycle. This was supported by our observation that introducing a pgi mutation into a metabolically rewired RuBiCO-dependent strain resulted in increased concentration of rPP cycle intermediates. This exemplifies the key role branch-points play in autocatalytic cycles (Gleizer et al. 2019; Gassler et al. 2020; Antonovsky et al. 2016; Barenholz et al. 2017; Flamholz et al. 2020; Herz et al. 2017), making them a leading target for rational attempts to integrate non-native autocatalytic cycles into metabolism.

The remaining two mutations act as global regulators, which presents a challenge in distinguishing their specific adaptive impacts from their broader effects on cellular expression patterns. Using proteomic analyses, we have managed to identify certain genes that had a major change in expression when one of the regulating mutations was introduced, and when both were introduced together.

One main observation was the down-regulation of several metabolic pathways including catabolism of alternative carbon sources and amino acid biosynthesis. We suggest this is an adaptive role in improving cell resource allocation, and a means to modulate resource allocation, possibly freeing up proteome resources from alternative carbon metabolism proteins. This includes proteins such as PTS system components or ATP dependent transporters, that tend to incur high energetic costs.

Other reasons these catabolic and transport pathways may result in reduced fitness could include competition over metabolites with the rPP cycle, or by causing other metabolic blunders such as dead end products with potential toxicity effects under the new flux regime. Another observed effect the crp mutation shows is the up-regulation of rpiB, a typically unexpressed isoenzyme to one of the reactions in the rPP cycle. This expression increase could be beneficial or even essential in a scenario where rpiA, the main enzyme catalysing this reaction, cannot support the flux required by the rPP cycle, for example due to kinetic capacity limitations or allosteric inhibition. Further work is needed to investigate if up-regulation of rpiB is indeed essential for the autotrophic phenotype.

We also suggest these mutations play a role in improving the utilisation of formate, which is supported by the up-regulation of fdoG & fdoH. We found that a strain introduced with both the crp and rpoB mutations demonstrated a pronounced up-regulation of over 4-fold and also a concrete fitness advantage in growth conditions which include formate. Metabolomics analysis showed this strain also has a significant increase in the NADH/NAD+ ratio. NADH is the essential cofactor for the electron transfer reaction of the non-native rPP cycle (catalysed by the D-glyceraldehyde-3P dehydrogenase GAPDH) and a product of the non-native formate-based energy module. The importance of this up-regulation is also supported by natural chemolithoautotrophic metabolism, which uses two distinct ways to harvest energy and reducing power from the electron-donor. For example, in the model chemolithoautotroph Cupriavidus necator, the oxidation of formate is catalysed by two distinct types of FDH: a soluble, NAD+-linked enzyme (S-FDH; EC 1.2.1.2) and a membrane-bound enzyme coupled directly to the respiratory chain (Friedrich, Bowien, and Friedrich 1979; Oh and Bowien 1998).

We suspect utilising formate dehydrogenase O (FDH-O) in the periplasm is enabling the cell to also utilise formate directly on the membrane to create a proton motive force for energy production without the need to exhaust the intracellular NADH pool, thus allowing for its accumulation. Increased levels of NADH creates favourable thermodynamics for the rPP cycle flux and could also offer improved kinetics. This result is in line with previous models, predicting that high cofactor levels are required for a stable metabolic state of the rPP cycle (Janasch et al. 2019). We suggest this observation offers an avenue for future rational design attempts.

We continue by hypothesising that such regulatory mutations are to be expected in adaptive laboratory evolution experiments, especially during selection for a major shift (novelty) in the phenotype. A major shift will often necessitate a modulation of the activity of many cellular components. Lab evolution is limited by population size and number of generations, which means it can only explore a small part of the genotypic space. Therefore, sets of three or more mutations, where each mutation on its own does not confer an advantage, have a negligible probability to occur. On the other hand, a single mutation in a global regulator could concurrently perturb the activity of hundreds of components and might be the only viable solution to capture several required changes simultaneously.

Another reported observation is how metabolic rewiring can serve as a “metabolic scaffold” to direct lab evolution. Since scaffold mutations are put in place for the purposes of selection, they can be removed once the goal is achieved. A common limitation of lab evolution is its tendency to follow evolutionary trajectories which increase fitness by deactivating or losing the heterologously expressed genes, thereby arriving at “dead-end” local optimum. A “metabolic scaffold” as utilised here can block these dead-end trajectories by coupling a target activity (here RuBisCO carboxylation) to growth, thereby increasing the likelihood of trajectories that lead to the desired phenotype. Once the phenotype is achieved, removal of the metabolic scaffold should be possible since growth is by definition dependent on the non-native genes. Using metabolic growth-coupling as a temporary “metabolic scaffold” that can be removed, could serve as a promising strategy for achieving minimally perturbed genotypes in future metabolic engineering attempts.

The approach presented here, of harnessing iterative lab evolution to find a compact set of essential mutations, strikingly shows the extreme malleability of metabolism and evolution capacity to switch trophic modes in laboratory time-scales with very few mutations. It can also serve as a promising pipeline for other genetic engineering attempts. Having a compact and well-defined genotype in a model organism allowed us to reveal metabolic adaptation steps needed for autotrophic growth, and could be used for future exploration of the design principles of the rPP cycle operation and its components (Aigner et al. 2017). The harnessing and rationalisation of the lab evolution process and metabolic adaptation steps elucidated here could facilitate the introduction of the rPP cycle and other carbon fixation cycles into other organisms of interest. Next steps could include integrating metabolic bio-production modules and alternative energy sources. Now that synthetic autotrophy in E. coli can be achieved with only a few defined steps, it expands the playground for more labs to join the quest for sustainable bio-production from CO2.

Supporting information

eLife Assessment

This is an important follow-up study to a previous paper in which the authors reconstituted CO2 metabolism (autotrophy) in Escherichia coli. Here, the authors define a set of just three mutations that promote autotrophy, highlighting the malleability of E. coli metabolism. The authors make a convincing case that mutations in pgi are loss-of-function mutations that prevent metabolic efflux from the reductive pentose phosphate autocatalytic cycle, and their data suggest possible roles of mutations in two other genes - crp and rpoB. This research will be particularly interesting to synthetic biologists, systems biologists, and metabolic engineers aiming to develop synthetic autotrophic microorganisms.

Joint Public Review:

The authors previously showed that expressing formate dehydrogenase, rubisco, carbonic anhydrase, and phosphoribulokinase in Escherichia coli, followed by experimental evolution, led to the generation of strains that can metabolise CO2. Using two rounds of experimental evolution, the authors identify mutations in three genes - pgi, rpoB, and crp - that allow cells to metabolise CO2 in their engineered strain background. The authors make a strong case that mutations in pgi are loss-of-function mutations that prevent metabolic efflux from the reductive pentose phosphate autocatalytic cycle. The authors also use proteomic analysis to probe the role of the mutations in crp and rpoB. While they do not reach strong conclusions about how these mutations promote autotrophic growth, they provide some clues, leading to valuable speculation.

Comments on revised version:

The authors have thoroughly addressed the reviewers' comments. The major addition to the paper is the proteomic analysis of single and double mutants of crp and rpoB. These new data provide clues as to the role of the crp and rpoB mutations in promoting autotrophic growth, which the authors discuss. The authors acknowledge that it will require additional experiments to determine whether the speculated mechanisms are correct. Nonetheless, the new data provide valuable new insight into the role of the crp and rpoB mutations. The authors have also expanded their description of the crp and rpoB mutations, making it clearer that the effects of these mutations are likely to be distinct, albeit with potential for overlap in function.

Author Response

The following is the authors’ response to the previous reviews.

Recommendations for the authors:

The single-mutant and double-mutant crp/rpoB strains were made by co-transduction with a nearby gene deletion (kanR-marked). I couldn't tell from the methods section whether these mutants, e.g., crp-H22N delta-chiA, were compared to wild-type cells or deletion mutants, e.g., delta chiA, in the proteomics experiments. I encourage the authors to explain this more clearly in the methods section, and to briefly mention in the Results section and relevant figure legends that the crp/rpoB mutant strains (and possibly the "wild-type" strains) also have gene deletions. If the comparison "wild-type" strains are fully wild-type (i.e., not deleted for chiA/yjaH), it is especially important to mention this in the Results section and the figure legends since the phenotypic changes could be due to the gene deletions rather than the mutations in crp/rpoB

We appreciate and agree with the editor's suggestion to clarify this point.

Accordingly, we have made the following changes to the text:

p11 L30-34 in the main text:

"The second experiment similarly compared an engineered BW25113 (BW) strain, containing the two regulatory mutations from the compact set (i.e., crp H22N and rpoB A1245V) together with the deletions used to insert them (see methods and DataS1 file), to a “wild type” BW strain (a corresponding knockout strain without the mutations, see methods)."

p28 under Chemostat proteomics experiment L13-16 in methods:

"The starting volume of each bioreactor was 150 ml M9 media supplemented with either 30 mM and 10mM D-xylose for the evolved and ancestor samples or only 10mM D-xylose for BW including compact set mutations and/or the deletions used for their insertions (DataS1 file). The minimal media also included trace elements and vitamin B1 was omitted."

Version 3

Abstract

Abstract

Synthetic autotrophy is a promising avenue to sustainable bioproduction from CO2. Here, we use iterative laboratory evolution to generate several distinct autotrophic strains. Utilising this genetic diversity, we identify that just three mutations are sufficient for E. coli to grow autotrophically, when introduced alongside non-native energy (formate dehydrogenase) and carbon-fixing (RuBisCO, phosphoribulokinase, carbonic anhydrase) modules. The mutated genes are involved in glycolysis (pgi), central-carbon regulation (crp), and RNA transcription (rpoB). The pgi mutation reduces the enzyme’s activity, thereby stabilising the carbon-fixing cycle by capping a major branching flux. For the other two mutations, we observe down-regulation of several metabolic pathways and increased expression of native genes associated with the carbon-fixing module (rpiB) and the energy module (fdoGH), as well as an increased ratio of NADH/NAD+ - the cycle’s electron-donor. This study demonstrates the malleability of metabolism and its capacity to switch trophic modes using only a small number of genetic changes and could facilitate transforming other heterotrophic organisms into autotrophs.

Introduction

Anthropogenic interference and intensive burning of fossil fuels releases large amounts of sequestered carbon in the form of CO2, a main driver of climate change (Olofsson and Hickler 2008; Friedlingstein et al. 2022). Direct air carbon capture, by any means, is urgently needed to help reduce emissions. Here, we focus on biology-based carbon fixation as a potential avenue for CO2 capture, which has been shown to be a potentially efficient production platform (Leger et al. 2021; Van Peteghem et al. 2022; Schwander et al. 2016). Specifically, we suggest using microbes for CO2 valorization and produce a diverse repertoire of products including food, fuel, bio-plastics or other commodities (Claassens et al. 2019; Gleizer et al. 2020; Li et al. 2012; Nielsen and Keasling 2016). Natural autotrophs already metabolise CO2 efficiently, yet most are challenging to cultivate and manipulate genetically compared to heterotrophic model organisms like E. coli. Here we explore an emerging alternative approach – converting heterotrophic model organisms into tractable autotrophs for bio-production (Gleizer et al. 2020; Baumschabl et al. 2022; Gassler et al. 2020; Gleizer et al. 2019). In addition to promising engineering applications, studying the transition from heterotrophic to autotrophic metabolism can uncover fundamental principles about the structure and regulation of carbon fixation metabolism (Gleizer et al. 2019; Gassler et al. 2020; Antonovsky et al. 2016; Barenholz et al. 2017; Flamholz et al. 2020; Herz et al. 2017).

Recently, with the increasing interest in platforms for sustainable production, synthetic biologists have been able to successfully manipulate different organisms to enable major metabolic transitions (Gassler et al. 2020; Gleizer et al. 2019; Kim et al. 2020; Chen et al. 2020; Keller et al. 2022; Bang et al. 2020; Satanowski et al. 2020; Yishai et al. 2018). These efforts resulted in the successful introduction of non-native C1 utilisation pathways, and even included successful transitions from heterotrophic into completely autotrophic metabolisms in bacteria (Gleizer et al. 2019) and yeast (Gassler et al. 2020). These previous studies introduced the reductive pentose phosphate (rPP) cycle, the most prominent carbon fixation pathway in nature (also known as the Calvin-Benson cycle), into Escherichia coli (E. coli) and Komagataella phaffii (previously Pichia pastoris). Indeed, we showed that E. coli was able to utilise the rPP cycle for the synthesis of all biomass carbon - thus converting it from a heterotroph to an autotroph (Gleizer et al. 2019). However, this transition required continuous culture in selective conditions for several months (adaptive lab evolution) and resulted in many uninterpretable mutations, exposing a large knowledge gap.

Using adaptive lab evolution presents a challenge as it is hard to isolate the changes necessary for a given phenotype from the overall set of accumulated mutations. Here we aimed to characterise the landscape of mutations required for autotrophic growth. Studying mutations that arise in multiple lineages can help us better understand the biological basis for the heterotroph-to-autotroph transition. To achieve this goal, we created a pipeline that enabled us to distinguish mutations essential for autotrophy from unspecific low-frequency mutations, i.e. ones that are generally promoted by the physiological conditions in evolution experiments. The final outcome of this pipeline is a compact essential set of changes permitting autotrophy in E. coli. We then proceeded with further analyses to shed light on the mechanism behind this compact set of essential mutations. We show the phenotype included a change in the activity of a central carbon metabolism enzyme and intracellular changes of the non-native carbon fixation cycle’s metabolite and co-factors pools. We conclude by speculating about the role of these metabolic changes.

Results

A compact set of mutations that enable autotrophic growth was identified using rational design, iterative lab evolution and genetic engineering

Adaptive lab evolution is an effective tool that enables integration of non-native metabolic pathways. In order to harness the power of adaptive lab evolution, a selection system must be implemented to direct the evolution towards the desired function. In our case, the selection for the activity of the non-native rPP cycle. To ensure that some level of carbon fixation by RuBisCO becomes essential for growth, we knocked out phosphofructokinase (pfkA and pfkB), and 6-phosphate-1-dehydrogenase (zwf) which creates a stoichiometric imbalance and growth arrest when growing on five carbon sugars such as xylose (Gleizer et al. 2019). This imbalance could be rescued by a metabolic bypass - the phosphorylation of ribulose-5-P by Prk and the subsequent carboxylation by RuBisCO, thereby coupling growth to the heterologously expressed rPP enzymes. Critically, this setup allowed us to evolve E. coli, a natural heterotroph, into an autotroph, using a non-native rPP cycle to utilise CO2 as a sole carbon-source. This system also requires formate dehydrogenase to oxidise formate for energy (Gleizer et al. 2019).

Although adaptive lab evolution is a powerful tool, it also presents a challenge - once a new phenotype has been achieved, it is hard to separate the essential changes required for its appearance from the overall set of accumulated mutations. In many cases, mutations occur on the background of highly adaptive ones, leading to a spread of neutral “hitchhiker” mutations in the population. In other cases, mutations that appeared at the beginning of the evolution are no longer needed for the final phenotype. Our goal was to find a set of mutations that, if introduced into a heterotrophic wild type E. coli, would result in the successful integration of the non-native carbon fixation and energy modules. This would enable autotrophic growth based solely on genetic engineering without the need for further evolutionary adaptation and potentially uncovering the mechanism behind this metabolic transition.

To isolate the essential genetic changes, we used a workflow that included three main stages as shown in Fig. 1. A) rational design-introduction of required heterologous genes and metabolic knockouts to enforce RuBisCO-dependent growth; B) adaptive lab evolution-revealing mutation candidates by growing the designed strain in autotrophic-selecting conditions over many generations until an autotrophic phenotype is obtained; C) evolution-inspired genetic engineering-introducing the most promising mutations revealed in stage B into the designed strain and testing it for autotrophic growth. Stages B and C were repeated until no further evolution was required, i.e. the designed strain in stage C had an autotrophic phenotype.

Autotrophic phenotype achieved by introducing 3 mutations on top of a rationally designed ancestor.

(A) We rationally designed the “wild-type” E. coli background strain (BW25113, depicted as a black bacterium) by introducing 4 enzymes: RuBisCO (cbbM), phosphoribulokinase (prkA), carbonic anhydrase (CA) and formate dehydrogenase (fdh), and 3 genomic knockouts: glucose-6-phosphate dehydrogenase (zwf) and phosphofructokinase A&B (pfkAB). We denote the resulting strain as “ancestor” (brown bacterium). (B) We tested the resulting strain for autotrophic growth and, since it didn’t grow, we used it for adaptive lab evolution in xylose-limited chemostats - i.e., conditions that select for higher carbon fixation flux. Altogether, we were able to isolate three evolved clones (two distinct strains from one chemostat experiment, Evolved I & II, and another strain from a second experiment, Evolved III, step B1). (C) Out of four consensus mutations that we identified, two - rpoB and pgi - were incorporated into the ancestor, giving “ancestor 2.0” (light blue bacterium, step C1). Once more, we tested for autotrophic growth and were unsuccessful. Therefore, we initiated another round of adaptive lab evolution experiments using ancestor 2.0 growing in two xylose limited chemostats and isolated two evolved clones (one from each chemostat, step B2). The two clones shared a single consensus mutation, in crp. We thus created a new strain from ancestor 2.0 by introducing the crp mutation into it (blue bacterium, step C2). This strain could grow autotrophically and thus we achieved a compact autotrophic strain.

To find the most promising mutations in stage B, we applied at least one of two criteria: mutations in genes that occur repeatedly in different autotrophic lineages (but are uncommon in other adaptive lab evolution experiments) or mutations that when reverted, result in a loss of phenotype. We discovered these “consensus” mutations by sequencing isolated autotrophic clones, obtained from repeated adaptive lab evolution experiments-similar to the conditions we used in Gleizer et al. 2019 (Fig. 1; steps B1, B2).

In practice, we inoculated the strain denoted as Ancestor into a chemostat with limited xylose and excess formate. After ≈3 months (≈60 chemostat generations) we isolated from the chemostat a clone able to grow under autotrophic conditions (methods), and compared its mutation set to those of two previously evolved autotrophic clones (Gleizer et al. 2019). The intersection of the three clones contained 4 genes that were mutated in all autotrophic isolated clones: the central carbon metabolism enzyme phosphoglucoisomerase (pgi), the beta subunit of the RNA polymerase (rpoB), poly A polymerase (pcnB) and the MalT DNA-binding transcriptional activator (malT) (Fig. 1, step B1). rpoB, pcnB and malT are genes that are commonly mutated in adaptive lab evolution experiments (Phaneuf, Gosting, and Palsson 2019), suggesting that they have little relevance for autotrophic growth. Therefore, we chose to focus on the pgi mutation.

We introduced the H386Y pgi mutation into the rationally designed ancestor (Fig. 1, step C1), and the engineered strain was tested for growth in autotrophic conditions (methods). No growth was observed in those conditions, which meant that additional mutations were necessary in order to achieve autotrophy.

Genomic sequencing revealed that during the genetic manipulation for introducing the pgi mutation, another mutation in rpoB (A1245V) appeared in the genome. In parallel, reverting a variety of RNA polymerase mutations in other evolved strains back to their wild-type allele, showed that mutations in RNA polymerase are in fact essential to the phenotype. Therefore, despite the fact that it was unintentional, and different from the other rpoB mutations observed in the ALE experiments, we left the mutation in the genome and proceeded with this strain.

This strain, mutated in both pgi and rpoB (denoted “ancestor 2.0”) was then used as the starting point for another round of evolution (Fig. 1, step B2). Since it had already contained two mutations that were likely to be relevant for the desired phenotype, we expected to observe evolved autotrophic strains after fewer generations, compared to the first round, and to find fewer mutations in them. Indeed, on two separate experiments, it took roughly one month until autotrophic growth was observed, as opposed to more than 90 days in previous attempts. Even though these strains (Fig. 1, “Evolved 4” and “Evolved 5”) both had multiple mutations, only one mutation was shared between these independent evolution experiments - a non-synonymous mutation in the cAMP receptor protein, crp (H22N). When we introduced this mutation to the “ancestor 2.0” background in the evolution-inspired genetic engineering stage (Fig. 1, step C2), the strain was immediately able to grow under autotrophic conditions.

Since the RNA polymerase mutation was introduced into this strain (rpoB A1245V) as a byproduct of genetic engineering and never observed in the adaptive evolution experiments, we needed to verify its essentiality to the autotrophic phenotype. We found that a reversion of rpoB A1245V back to the wild-type allele causes a loss of the autotrophic phenotype. We therefore could conclude that three mutations (pgi*, crp* and rpoB*) are sufficient to facilitate the autotrophic phenotype on top of the heterotrophic ancestor background. We denote the final engineered autotrophic E. coli strain, the “compact autotrophic E. coli”.

We verified the genotype using whole genome sequencing (methods). The results included the rationally designed knockouts (ΔpfkA, ΔpfkB, Δzwf), heterologous plasmids (energy and carbon-fixing modules) and the three introduced mutations (pgi*, crp* and rpoB*). During the genetic engineering process, two additional mutations occurred unintentionally in the genes uhpT and yejG. uhpT is a hexose transporter and is unlikely to have any effect on the autotrophic phenotype, especially because the mutation was an early stop codon (Q7*) and likely a loss of function. As the function of yejG is unknown, we wanted to ensure that it is not essential to the autotrophic phenotype. Therefore, we reverted the mutation back to its wild-type allele and found that indeed the cells were still autotrophic.

Thus, we constructed a new autotrophic strain with three essential mutations (pgi, crp & rpoB), which we denoted the compact strain. When comparing this compact strain to a previously characterised autotrophic strain (Gleizer et al. 2019), it was nearly identical in terms of growth rate, lag time and yield (Fig. 2A). Using 13C labelling, we verified that all the biomass carbon stems from CO2 (Fig. 2B). We termed the three introduced mutations (pgi, crp & rpoB) together with the carbon fixation machinery (cbbM, prkA & CA) and the energy module (fdh) the “autotrophic enabling gene set”.

Validation and characterisation of the autotrophic phenotype.

(A) left: growth curve of an isolated evolved clone vs the engineered compact strain in liquid M9 minimal media supplemented with 45mM sodium formate and sparged with a gas mixture of 10% CO2, 5% oxygen, and 85% nitrogen. right: calculated growth rate using linear regression of the engineered compact strain (light blue) and evolved strain (orange), the calculated doubling time with the given conditions is about 24h (μ≈0.03h-1). Growth was carried out in triplicates (n=3) in a Spark plate reader with gas control, dark grey area corresponds to the standard deviation of the mean (150 μl culture + 50 μl mineral oil, to prevent evaporation). (B) The fractional contribution of 13CO2 to various protein-bound amino acids of the compact autotrophic strain after ≈4 doublings on 13CO2 and 13C labelled formate (light blue, mean of n=3). This reached the expected 13C labelling fraction (see methods) of the biosynthesized amino acids. When grown in naturally labelled CO2 and 13C labelled formate (dark blue, n=1, technical triplicate) the 13C content dropped to close to natural abundance. Experiments with 13CO2 as the substrate were carried out in air-tight (i.e., sealed) growth vessels.

Introduction of autotrophy-enabling gene set into wild-type background does not require bypass-prevention for growth

As described above, in order to select for autotrophic growth, the native E. coli was metabolically re-wired by 3 auxiliary genomic knockouts (ΔpfkA, ΔpfkB, Δzwf). These knockouts created a dependency on carboxylation by RuBisCO for growth even when consuming pentose sugars which were supplemented during the chemostat evolution (see methods, Antonovsky et al. 2016; Gleizer et al. 2019). This dependency was expected to direct the evolution towards increased usage of the synthetic rPP cycle and, eventually, to autotrophic growth.

Once autotrophic growth was achieved, and the enabling gene set was found, we wanted to test if these auxiliary knockouts (pfkA pfkB and zwf) were still required for the phenotype. We introduced the set of autotrophy-enabling genes into a wild-type (BW25113) E. coli background (methods, Fig. 3). After doing so, we discovered an unintentional genomic mutation, in the ribosomal 16S subunit (rrsA) that was introduced during the genomic manipulation process. The engineered E. coli strain was able to grow autotrophically, although its final OD was half and its growth rate was less than half of that of the rationally designed compact strain still harbouring the auxiliary knockouts (Fig. S2). This indicates that the “autotrophic enabling gene set” is sufficient to achieve the trophic mode change and that the knockouts are helpful but not essential for the final autotrophic growth. As discussed below, this demonstrates that, in some cases, metabolic rewiring could be viewed as a temporary “metabolic scaffold” which can be removed from the genotype once the desired phenotype is achieved.

Auxiliary knockouts are not required for final phenotype.

We transformed a wild-type BW25113 E. coli strain (black bacteria) with the carboxylating and energy module plasmids (grey circles with coloured gene annotations), and inserted 3 auxiliary genomic knockouts (red octagon) to rewire metabolism toward carboxylating dependency. This strain was used for iterative evolution experiments in order to generate diverse autotrophic strains and reveal mutations candidates for rational design, i.e. consensus mutations. The identified autotrophic enabling mutations (thin grey circle) were then introduced in a wild type strain expressing the heterologous plasmids but without the auxiliary knockouts, the final strain was able to grow in autotrophic conditions. Dashed lines represent gene/mutation introduction.

We thus established a new engineered autotrophic E. coli strain containing a set of three essential and sufficient mutations, and showed that the bypass-prevention knockouts introduced before the ALE are themselves not necessary for the phenotype. Due to the small required number of genetic modifications, this strain provides an opportunity to derive guidelines for future engineering efforts that use the rPP cycle. Having this goal in mind, we focused our attention on identifying the relevant phenotypes of each of these mutations and explaining their adaptive advantage. Because the mutations were essential for the autotrophic growth, we could not use autotrophic conditions to compare them to a strain with wild-type alleles. Therefore, we chose the most suitable combinations of strains and conditions that could isolate the effect of the pertinent mutation.

Modulation of a metabolic branch-point activity increased the concentration of rPP metabolites

Metabolic pathways that regenerate and synthesise more of their own metabolites are referred to as autocatalytic cycles. Within the autotrophic E. coli the rPP cycle is autocatalytic. Due to the inherent positive feedback mechanism, autocatalytic cycles tend to be unstable, and therefore the fluxes of entry and exit points (bifurcation/branch points) need to be balanced (Barenholz et al. 2017). Such tuning is needed since any disruption of the balance between cycling and branching fluxes could result in the depletion, or alternatively toxic accumulation of intermediate metabolites, and cycle arrest.

We previously predicted that mutations in branch points will be needed to stabilise the steady state flux within the rPP cycle (Barenholz et al. 2017). In line with this prediction we find that the mutation in phosphoglucoisomerase (Pgi), a key branch point in the rPP cycle, follows this design principle. Pgi consumes fructose-6-phosphate (F6P), one of the cycle intermediates, converting it into glucose-6-phosphate (G6P), a precursor for cell membrane biosynthesis. Thus, Pgi regulates flux out of the autocatalytic cycle (Fig. 4A).

Pgi mutation is required to partition flux towards regeneration of autocatalytic cycle substrate.

(A) Metabolic scheme of rPP cycle components when growing on xylose. (B) In vitro spectrophotometric coupled assay determined that the rate of the isomerization of fructose-6-phosphate (F6P) to glucose-6-phosphate (G6P) is ≈100-fold lower for purified PgiH386Y (≈1.5 [s-1]) compared to Pgiwild-type (≈130 [s-1]), n=3. Noise level was determined by negative control samples containing a non-Pgi enzyme. (C) Left: The wild type pgi, competes for its substrate F6P with the autocatalytic cycle, resulting in low F6P pool and low regeneration rate of ribulose-5-phosphate (Ru5P), which means that the RuBisCO-dependent pathway requires constant xylose supply. Right: The mutated Pgi (green) has reduced activity, which increases the ratio between the flux in the cycle and efflux which is needed for a stable regenerative flux towards Ru5P (via the pentose-phosphate-pathway), thus enabling an autotrophic cycle. (D) Left: Measured relative intracellular ratio of F6P and G6P of the ancestor background harbouring a PgiH386Y mutation compared to the ancestor with a wild-type Pgi, both grown in RuBisCO dependent conditions, n=6 cultures p-value < 0.05. Right: Relative intracellular abundance of pentose-phosphate-pathway metabolites - sedoheptulose-7P (S7P) and total pool of pentose phosphates (ribulose-5P, ribose-5P, ribose-1P, and xylulose-5P - denoted P5P) in a PgiH386Y strain versus the Pgiwild-type strain, both growing in a RuBisCO-dependent manner, n=3 cultures.

In the different adaptive lab evolution experiments, we found three distinct mutations in the pgi gene. The first, H386Y, is a non-synonymous mutation occurring in one of the catalytic residues in the active site, and is part of the autotrophic enabling gene set. The second was a complete knockout, a 22KB chromosomal deletion, including also 16 other genes. The third was an early stop codon E72*. These observations led us to suspect the H386Y mutation in pgi decreases or even completely eliminates the activity of the enzyme.

We tested the kinetic rates of the isomerization of F6P to G6P of two purified Pgi enzymes, wild type Pgi and Pgi H386Y, using a spectroscopic coupled assay, where G6P was coupled to NADP+ reduction (Widjaja et al. 1998) (Fig. 4B; see also methods). The measured kcat of the wild type Pgi was ≈130 [s-1] (±26) which is in line with previous computational predictions of a kcat of ≈200 [s-1](Davidi et al. 2016). The mutated Pgi showed weak activity with a kcat of ≈1.5 [s-1](±1). Therefore the pgi mutation is in line with the original expectation that branching point regulation is necessary, and the nature of the observed mutations in pgi supports the notion that their role is to reduce the efflux from the non-native autocatalytic cycle.

Pgi is part of glycolysis/gluconeogenesis, where significant flux is required in wild type E. coli. Accordingly, wild type Pgi catalyses a reaction that is causing a strong efflux of F6P from the cycle, diminishing the regeneration of ribulose-1,5-bisphosphate (RuBP) which is usually not needed in the native metabolism but is essential for the rPP autocatalytic activity. In the mutated version of Pgi the efflux capacity is diminished, which can stabilise the cycle (Fig. 4C).

Following the same logic, we can expect an increase in the F6P substrate metabolite pool due to reduced consumption by Pgi. Therefore, we measured intracellular sugar-phosphates in the pgi mutant strain and a strain with a wild-type pgi allele by liquid chromatography-tandem mass spectrometry (LC-MS/MS). For this comparison, we used the RuBisCO-dependent ancestor strain as the genetic background. We found that the ratio of F6P to G6P was about 3 times higher in the pgi mutant strain relative to the wild-type (Fig. 4D). Furthermore, the pgi mutant had higher levels of metabolites within the rPP cycle (Fig. 4D), confirming the stabilising function of the mutation.

These results lead us to conclude that the recurring mutations in the pgi gene are mediating the integration of the non-native carbon fixation genes into a stable autocatalytic cycle. This is achieved by reducing the ratio of the efflux from F6P to G6P relative to the flux in the cycle, thereby redirecting it towards the rPP cycle and the regeneration of RuBP.

Proteomic analysis reveals up-regulation of rPP cycle and formate-associated genes alongside down-regulation of catabolic genes

The other two mutations in the compact set - crp (H22N) and rpoB (A1245V), are global effectors in the cell. Crp is one of the most general transcription factors in the cell and has a role in many processes from metabolism to osmoregulation and biofilm formation (Landis, Xu, and Johnson 1999; Jackson, Simecka, and Romeo 2002; Ammar, Wang, and Rao 2018). The position H22 in Crp is part of “activation region 2” of the protein, an interaction region with RNA polymerase (Fig. S8). This position was shown to have an effect on promoter specificity (Niu et al 1996) and was previously reported to have a role in the integration of the rPP cycle in E. coli (Herz et al. 2017).

Similarly, mutations in rpoB, a subunit of RNA polymerase, have been shown to have a large effect on the cell transcriptome (Conrad et al. 2010; Utrilla et al. 2016). The observed A1245V mutation in rpoB is situated within the core of RNA polymerase, proximal to its interaction with the DNA (figure S8). Furthermore, this genetic variation was previously documented in the adaptation of the E. coli B REL1206 strain to high temperatures (Tenaillon et al. 2012) and is closely positioned to another previously reported rpoB mutation (K1242Q) that appeared in adaptation to loss of Pgi activity (Charusanti et al. 2010).

Depending on the transcribed gene target, the effect of the two mutations might be additive, antagonistic, or synergistic. Since each one of these mutations individually (in combination with the pgi mutation) is not sufficient to achieve autotrophic growth, it is reasonable to assume that only the target genes whose levels of expression change significantly in the double-mutant are the ones relevant for the autotrophic phenotype. Given the accumulated knowledge described above, these mutations are likely to have a broad effect on the cell. We chose to further elucidate their effect using a proteomics approach. As such, we conducted two experiments in an attempt to find changes in expression that are relevant for the adaptation to the new trophic mode. The first experiment compared the proteome of an evolved clone to its ancestor. The second experiment similarly compared an engineered BW25113 (BW) strain, containing the two regulatory mutations from the compact set (i.e., crp H22N and rpoB A1245V) together with the deletions used to insert them (see methods and DataS1 file), to a “wild type” BW strain (a corresponding knockout strain without the mutations, see methods).

Since our aim was to observe only direct effects induced by the genetic changes, rather than by the environment, we measured the proteomes of the strains under identical conditions and dilution rates. Ideally, the comparison should be done in conditions that require autotrophic growth (since some effects might depend on the lack of an organic carbon source), however, all strains except the evolved clone cannot grow in such conditions. Therefore, we decided on growing the cells in the conditions present at the beginning of the lab evolution experiment, namely a xylose-limited chemostat. The dilution rate was set to ≈0.03 h-1 leading to a doubling time of ≈24h (methods).

In order to identify the genes with a large and significant fold change (fold change >4, p-value <0.05) we generated a volcano plot for each such comparison (Fig. S6). We then proceeded by comparing the output of the first experiment (evolved divided by ancestor, Fig. S6A) to that of the double mutant divided by the base BW strain (Fig. S6B), since both mutations are essential for the phenotype and are known to affect expression. We focused on genes with common trends of expression level in both sets (Fig. 5).

Proteomics analyses of fold change in evolved/ancestor expression levels versus double mutant/WT.

Comparison of the log2 fold change in the evolved autotrophic strain divided by the ancestor expression versus BW rpoB A1245V & crp H22N, divided by the WT. The size of dots represents significance, only genes with >4 fold-change who showed significance (<0.05 p-value) were annotated.

The proteomics analysis yielded few genes that had an upregulation of >4-fold in both the evolved versus ancestor and the double mutant versus BW base strain experiments. Two of these up-regulated genes, fdoG and fdoH, are subunits of the native formate dehydrogenase O enzyme complex (FDH-O) and are therefore related to the energy module. These genes were upregulated in both the evolved strain and in the strain with the crp H22N and rpoB A1245V double mutation (≈5-fold increase in both cases). The FDH-O enzyme complex takes part in the respiratory chain formed by formate dehydrogenase and dimethyl sulfoxide (DMSO) reductase. Here, the transfer of electrons from formate to DMSO (via a quinone electron carrier) is coupled to the generation of a proton-motive force across the cytoplasmic membrane for energy formation.

Another observed upregulated protein was ribose-phosphate-isomerase B (rpiB), which was up-regulated in the evolved strain (>40-fold increase) and in the double mutant ≈5-fold increase (Fig. 5). rpiB is an inducible isoenzyme of the constitutively expressed ribose-phosphate-isomerase A, a core enzyme in the rPP cycle. This might serve as evidence for a higher flux requirement through the pentose-phosphate-pathway, supported by expressing a second isoenzyme. Other explanations may include some form of regulation, perhaps allosteric, inhibiting RpiA activity, making the strain dependent on RpiB activity, or non-specific enzymatic activity of RpiB which is required by the cell. The fourth up-regulated gene - ftsI, a peptidoglycan DD-transpeptidase - does not have a clear explanation at this time.

At the other end of the expression spectrum, several pathways were down-regulated in both the evolved and in the BW double mutant strain (Fig. 5). Down-regulation of the starvation response, metabolic and PTS component genes seems to be at the heart of the adaptation to autotrophic growth, with several operons experiencing >4-fold decrease in expression. Most notably the gat, ara and man operons responsible for galactitol, arabinose and mannose degradation, respectively. Again, this is evident in both the evolved and the double mutant strains (Fig. 5). Other highly down-regulated metabolic pathways include amino acid biosynthesis, specifically valine and leucine, moreover we observed a down regulation in rspA & rspB, which encode for starvation sensing proteins. We suggest these results mean that many pathways usually up-regulated in E. coli during starvation may be redundant or even counter-productive in an autotrophic context, and decreasing their expression could result in increased fitness.

We additionally tested whether separating between the two regulatory mutations (crp and rpoB) might provide a clearer picture, e.g. if different up- or down-regulated genes are targeted by only one of these mutations and not the other. Therefore, we used single-mutant BW strains and compared them to the same BW wild-type as before. Essentially, the results did not change our conclusions and although in most cases the effect of the two mutations was cumulative, we did find at least one case where they were antagonistic (Fig. S7). For example, fdoG and fdoH showed a ≈4-fold decrease compared to wild type when the crp mutation was introduced, but (as stated earlier) had a ≈5-fold abundance increase in the double mutant, suggesting an epistatic effect between the rpoB and crp mutations (Fig. S6B & Fig. S7B). All the comparisons based on the single and double mutant strains can be found in the supplementary.

Mutations in regulatory genes lead to increased availability of the carbon fixation cycle electron donor NADH

Along with the effect on expression, we suspected the two transcription-associated mutations (crp & rpoB) to also have a metabolic and physiological effect. Since Crp and RpoB are both essential to the autotrophic phenotype and are known to indirectly interact in the cell (Savery, Rhodius, and Busby 1996; Niu et al. 1996; Frendorf et al. 2019), we address them as one unit. The proteomics analysis, where the strain with both mutations seem to have a mutual effect that is in some cases different than in the individual mutants, supported this decision. Mutations in crp and rpoB are known to affect growth at different conditions such as types of media or carbon sources (Utrilla et al. 2016; Conrad et al. 2010; Cheng et al. 2014; Frendorf et al. 2019). Therefore we tested whether this was also the case for the conditions under which the cells evolved in the chemostat.

We performed a growth experiment of the double mutant (crp H22N, rpoB A1245V) expressing fdh in a minimal medium containing xylose and formate. Under such conditions, the strain harbouring both mutations and fdh exhibited >25% increase in the final OD compared to the wild type expressing fdh (Fig. 6A&B). At the same time they also showed a longer (>10 hours) lag time. Notably, when grown solely on xylose, in the absence of formate, the increase in yield was not observed, and the extended lag was less pronounced (Fig. 6A).

mutations in rpoB and crp increase yield and intracellular NADH/NAD+ levels in fdh-expressing E. coli in the presence of formate.

Growth experiment of BW25113 wild-type (grey) compared to a crp H22N rpoB A1245V mutant (orange). Both strains express fdh. (A) strains were grown in 1g/L xylose and 40mM formate (solid line) or in the absence of formate (dashed line). The experiment was executed in a 96 well plate in 10% CO2 atmosphere, n=6 repeats. Lines represent the mean, light-grey background represents the standard deviation. (B) Maximal OD600 of the wild-type (grey) and mutant (orange) strains grown in 1g/L xylose and 40mM formate or in the absence of formate. Bar heights represent averages (n=6) of the median of the top 10 OD measurements of each replicate. Error bars represent standard error. (C) Intracellular NADH/NAD+ ratio of the wild-type (grey) was compared to the mutant (orange). The strains were grown in 2g/L xylose and 30mM formate, or in the absence of formate. The y-values are NADH/NAD+ ratios as fold-changes relative to wild-type without formate. Boxes represent 25-75 percentile ranges and dark lines represent median values. All data points are depicted without removing outliers.

Since NADH is the electron donor for carbon fixation, and a product of formate consumption via Fdh, we used metabolomics to measure the ratio of NADH/NAD+ in these strains in similar conditions (methods). All of the measured ratios were normalised to the wild-type (with fdh) growing in the absence of formate. The addition of 30mM formate, which can serve as an electron donor via oxidation to CO2 by Fdh, increased NADH/NAD+ ratio by >20-fold in the double mutant, significantly higher than the ≈10-fold increase in the wild-type strain (p-value <0.05). No significant difference was observed between the strains in the absence of formate (Fig. 6C).

NADH could affect growth in several ways, as discussed below. The specific reason for the increase in final OD or lag time remains inconclusive at this point but we hypothesise that it is related to the increased expression of Fdo observed in the proteomic analysis. We also note that as the proteomics results suggest - the mutations in global transcriptional machinery like Crp & RpoB could have many other effects on the autotrophic phenotype (Haverkorn van Rijsewijk et al. 2011) which await future exploration.

Discussion

In this study, we demonstrated how iterative evolution and engineering were able to narrow down the genomic mutations required for autotrophy in E. coli from tens of mutations to only three mutated genes. The three mutated genes - pgi, crp and rpoB - serve as a compact set, which when introduced into a wild-type background and supplemented with the carbon-fixing and energy modules, achieve an autotrophic phenotype. By metabolic, proteomic and physiological characterisation we were able to distil some of the effects of these mutations and understand their role in facilitating the autotrophic phenotype. We suggest that the three main effects achieved by the autotrophy-enabling mutations are: (i) integrating the non-native carbon fixation cycle by enhancing flux within the cycle over outgoing flux; (ii) facilitating the integration of the energy module via decoupling of formate harvesting into energy and redox, thereby increasing the NADH/NAD+ ratio, which could have both thermodynamic and kinetic benefits (iii) and freeing protein resources by decreasing unneeded expression.

The mutation in pgi affects the activity of its gene product phosphoglucoisomerase - a metabolic enzyme that catalyses a reaction that branches out of the non-native rPP cycle. The enzyme converts F6P (a cycle intermediate) to G6P (a precursor required for membrane biosynthesis). In a previous study, this mutation was found to be essential for the integration of the rPP autocatalytic cycle and was part of a minimal set for hemi-autotrophic growth (Gleizer et al. 2019; Gassler et al. 2020; Antonovsky et al. 2016; Barenholz et al. 2017; Flamholz et al. 2020; Herz et al. 2017). As our results suggest, the mutated enzyme shows a significant reduction in activity, which decreases the efflux from the rPP cycle to biosynthesis. We hypothesised that its role is to generate a new, higher, steady-state concentration for the rPP cycle metabolites, thus stabilising the cycle. This was supported by our observation that introducing a pgi mutation into a metabolically rewired RuBiCO-dependent strain resulted in increased concentration of rPP cycle intermediates. This exemplifies the key role branch-points play in autocatalytic cycles (Gleizer et al. 2019; Gassler et al. 2020; Antonovsky et al. 2016; Barenholz et al. 2017; Flamholz et al. 2020; Herz et al. 2017), making them a leading target for rational attempts to integrate non-native autocatalytic cycles into metabolism.

The remaining two mutations act as global regulators, which presents a challenge in distinguishing their specific adaptive impacts from their broader effects on cellular expression patterns. Using proteomic analyses, we have managed to identify certain genes that had a major change in expression when one of the regulating mutations was introduced, and when both were introduced together.

One main observation was the down-regulation of several metabolic pathways including catabolism of alternative carbon sources and amino acid biosynthesis. We suggest this is an adaptive role in improving cell resource allocation, and a means to modulate resource allocation, possibly freeing up proteome resources from alternative carbon metabolism proteins. This includes proteins such as PTS system components or ATP dependent transporters, that tend to incur high energetic costs.

Other reasons these catabolic and transport pathways may result in reduced fitness could include competition over metabolites with the rPP cycle, or by causing other metabolic blunders such as dead end products with potential toxicity effects under the new flux regime. Another observed effect the crp mutation shows is the up-regulation of rpiB, a typically unexpressed isoenzyme to one of the reactions in the rPP cycle. This expression increase could be beneficial or even essential in a scenario where rpiA, the main enzyme catalysing this reaction, cannot support the flux required by the rPP cycle, for example due to kinetic capacity limitations or allosteric inhibition. Further work is needed to investigate if up-regulation of rpiB is indeed essential for the autotrophic phenotype.

We also suggest these mutations play a role in improving the utilisation of formate, which is supported by the up-regulation of fdoG & fdoH. We found that a strain introduced with both the crp and rpoB mutations demonstrated a pronounced up-regulation of over 4-fold and also a concrete fitness advantage in growth conditions which include formate. Metabolomics analysis showed this strain also has a significant increase in the NADH/NAD+ ratio. NADH is the essential cofactor for the electron transfer reaction of the non-native rPP cycle (catalysed by the D-glyceraldehyde-3P dehydrogenase GAPDH) and a product of the non-native formate-based energy module. The importance of this up-regulation is also supported by natural chemolithoautotrophic metabolism, which uses two distinct ways to harvest energy and reducing power from the electron-donor. For example, in the model chemolithoautotroph Cupriavidus necator, the oxidation of formate is catalysed by two distinct types of FDH: a soluble, NAD+-linked enzyme (S-FDH; EC 1.2.1.2) and a membrane-bound enzyme coupled directly to the respiratory chain (Friedrich, Bowien, and Friedrich 1979; Oh and Bowien 1998).

We suspect utilising formate dehydrogenase O (FDH-O) in the periplasm is enabling the cell to also utilise formate directly on the membrane to create a proton motive force for energy production without the need to exhaust the intracellular NADH pool, thus allowing for its accumulation. Increased levels of NADH creates favourable thermodynamics for the rPP cycle flux and could also offer improved kinetics. This result is in line with previous models, predicting that high cofactor levels are required for a stable metabolic state of the rPP cycle (Janasch et al. 2019). We suggest this observation offers an avenue for future rational design attempts.

We continue by hypothesising that such regulatory mutations are to be expected in adaptive laboratory evolution experiments, especially during selection for a major shift (novelty) in the phenotype. A major shift will often necessitate a modulation of the activity of many cellular components. Lab evolution is limited by population size and number of generations, which means it can only explore a small part of the genotypic space. Therefore, sets of three or more mutations, where each mutation on its own does not confer an advantage, have a negligible probability to occur. On the other hand, a single mutation in a global regulator could concurrently perturb the activity of hundreds of components and might be the only viable solution to capture several required changes simultaneously.

Another reported observation is how metabolic rewiring can serve as a “metabolic scaffold” to direct lab evolution. Since scaffold mutations are put in place for the purposes of selection, they can be removed once the goal is achieved. A common limitation of lab evolution is its tendency to follow evolutionary trajectories which increase fitness by deactivating or losing the heterologously expressed genes, thereby arriving at “dead-end” local optimum. A “metabolic scaffold” as utilised here can block these dead-end trajectories by coupling a target activity (here RuBisCO carboxylation) to growth, thereby increasing the likelihood of trajectories that lead to the desired phenotype. Once the phenotype is achieved, removal of the metabolic scaffold should be possible since growth is by definition dependent on the non-native genes. Using metabolic growth-coupling as a temporary “metabolic scaffold” that can be removed, could serve as a promising strategy for achieving minimally perturbed genotypes in future metabolic engineering attempts.

The approach presented here, of harnessing iterative lab evolution to find a compact set of essential mutations, strikingly shows the extreme malleability of metabolism and evolution capacity to switch trophic modes in laboratory time-scales with very few mutations. It can also serve as a promising pipeline for other genetic engineering attempts. Having a compact and well-defined genotype in a model organism allowed us to reveal metabolic adaptation steps needed for autotrophic growth, and could be used for future exploration of the design principles of the rPP cycle operation and its components (Aigner et al. 2017). The harnessing and rationalisation of the lab evolution process and metabolic adaptation steps elucidated here could facilitate the introduction of the rPP cycle and other carbon fixation cycles into other organisms of interest. Next steps could include integrating metabolic bio-production modules and alternative energy sources. Now that synthetic autotrophy in E. coli can be achieved with only a few defined steps, it expands the playground for more labs to join the quest for sustainable bio-production from CO2.

Supporting information